---
title: "irtQ"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

The goal of `irtQ` is to fit unidimensional item response theory (IRT) models to data that may include both dichotomous and polytomous items. The package enables:

- Typical item parameters estimation 
- Pretest item calibration
- Multiple-group item calibration
- Estimation of examinees' latent abilities,
- Evaluation of model-data fit at the item level

Item parameter estimation is conducted using marginal maximum likelihood estimation via the expectation-maximization (MMLE-EM) algorithm (Bock & Aitkin, 1981).  
For pretest item calibration, `irtQ` supports:

- Fixed item parameter calibration (FIPC; Kim, 2006),
- Fixed ability parameter calibration (FAPC; Ban et al., 2001; Stocking, 1988).

For ability estimation, several widely used scoring methods are available, including:

- Maximum likelihood estimation (ML)
- Maximum likelihood estimation with fences (MLF; Han, 2016)
- Weighted likelihood estimation (WL; Warm, 1989)
- Maximum a posteriori estimation (MAP; Hambleton et al., 1991)
- Expected a posteriori estimation (EAP; Bock & Mislevy, 1982)
- EAP summed scoring (Thissen et al., 1995; Thissen & Orlando, 2001)
- Inverse test characteristic curve (TCC) scoring (e.g., Kolen & Brennan, 2004; Kolen & Tong, 2010; Stocking, 1996)

Also, model fit assessment includes item fit statistics such as:

- Chi-square (X²; Bock, 1960; Yen, 1981),
- Likelihood ratio chi-square (G²; McKinley & Mills, 1985),
- Infit and outfit statistics (Ames et al., 2015),
- Graphical residual diagnostics (Hambleton et al., 1991).
- S-X² (Orlando & Thissen, 2000, 2003),  

In addition, the package offers a variety of utilities for IRT analysis, including:

- Detecting DIF (e.g., RDIF, CATSIB)
- Computing classification accuracy and consistency indices
- Simulating response data
- Computing the conditional distribution of observed scores using the Lord-Wingersky recursion
- Calculating item and test information and characteristic functions
- Visualizing item and test characteristic and information curves
- Importing item or ability parameters from popular IRT software
- Running flexMIRT (Cai, 2017) directly from R
- Supporting additional tools for flexible and practical IRT analyses

## Installation

You can install the released version of irtQ from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("irtQ")
```

You can also install the latest development version of `irtQ` from GitHub using the **devtools** package:

```r
install.packages("devtools")
devtools::install_github("hwangQ/irtQ")
```

## 1. Item Calibration for a Linear Test Form

Item parameter estimation for a linear test form can be performed using the `irtQ::est_irt()` function, which implements marginal maximum likelihood estimation via the expectation-maximization (MMLE-EM) algorithm (Bock & Aitkin, 1981). The function returns item parameter estimates along with their standard errors, computed using the cross-product approximation method (Meilijson, 1989).

The `irtQ` package supports calibration for mixed-format tests containing both dichotomous and polytomous items. It also provides a flexible set of options to address various practical calibration needs. For example, users can:

- Specify prior distributions for item parameters
- Fix specific parameters (e.g., the guessing parameter in the 3PL model)
- Estimate the latent ability distribution using a nonparametric histogram method (Woods, 2007)

In the `irtQ` package, item calibration for a linear test form typically involves two main steps:

1. Prepare the examinees' response data set for the linear test form

     To estimate item parameters using the `irtQ::est_irt()` function,
     a response data set for the linear test form must first be prepared.
     The data should be provided in either a matrix or data frame format,
     where rows represent examinees and columns represent items. If there
     are missing responses, they should be properly coded (e.g., `NA`).

2. Estimate item parameters using the `irtQ::est_irt()` function

     To estimate item parameters, several key input arguments must be
     specified in the `irtQ::est_irt()` function:
     - `data`: A matrix or data frame containing examinees' item responses.
     - `model`: A character vector specifying the IRT model for each item
        (e.g., `"1PLM"`, `"2PLM"`, `"3PLM"`, `"GRM"`, `"GPCM"`).
     - `cats`: A numeric vector indicating the number of score categories for
        each item. For dichotomous items, use 2.
     - `D`: A scaling constant (typically 1.702) to align the logistic
        function with the normal ogive model.

     Optionally, you may incorporate prior distributions for item parameters:
     - `use.aprior`, `use.bprior`, `use.gprior`: Logical indicators specifying
        whether to apply prior distributions to the discrimination (`a`),
        difficulty (`b`), and guessing (`g`) parameters, respectively.
     - `aprior`, `bprior`, `gprior`: Lists specifying the distributional form
        and corresponding parameters for each prior. Supported distributions
        include Beta, Log-normal, and Normal.

     If the response data contain missing values, you must specify the missing
     value code via the `missing` argument.

     By default, the latent ability distribution is assumed to follow a standard
     normal distribution (i.e., N(0, 1)). However, users can estimate the empirical
     histogram of the latent distribution by setting `EmpHist = TRUE`, based on the
     nonparametric method proposed by Woods (2007).


## 2. Pretest Item Calibration with the Fixed Item Parameter Calibration (FIPC) Method (e.g., Kim, 2006)

The fixed item parameter calibration (FIPC) method is a widely used approach for calibrating pretest items in computerized adaptive testing (CAT). It enables the placement of parameter estimates for newly developed items onto the same scale as the operational item parameters (i.e., the scale of the item bank), without the need for post hoc linking or rescaling procedures (Ban et al., 2001; Chen & Wang, 2016).

In FIPC, the parameters of the operational items are fixed, and the prior distribution of the latent ability variable is estimated during the calibration process. This estimated prior is used to place the pretest item parameters on the same scale as the fixed operational items (Kim, 2006).

In the `irtQ` package, FIPC is implemented through the following three steps:

1. Prepare the item metadata, including both the operational items (to be fixed) and the pretest items.

     To perform FIPC using the `irtQ::est_irt()` function, the item metadata must
     first be prepared. The item metadata is a structured data frame that includes
     essential information for each item, such as the number of score categories
     and the IRT model type. For more details, refer to the **Details** section
     of the `irtQ::est_irt()` documentation.

     In the FIPC procedure, the metadata must contain both:
     - Operational items (whose parameters will be fixed), and
     - Pretest items (whose parameters will be freely estimated).

     For the pretest items, the `cats` (number of score categories) and `model`
     (IRT model type) must be accurately specified. However, the item parameter
     values (e.g., `par.1`, `par.2`, `par.3`) in the metadata serve only as
     placeholders and can be arbitrary, since the actual parameter estimates
     will be obtained during calibration.

     To facilitate creation of the metadata for FIPC, the helper function
     `irtQ::shape_df_fipc()` can be used.

 2. Prepare the response data set from examinees who answered both the operational
 and pretest items.

     To implement FIPC using the `irtQ::est_irt()` function, examinees' response
     data for the test form must be provided, including both operational and
     pretest items. The response data should be in a matrix or data frame format,
     where rows represent examinees and columns represent items.
     Note that the column order of the response data must exactly match
     the row order of the item metadata.

 3. Perform FIPC using the `irtQ::est_irt()` function to calibrate the pretest items.

     When FIPC is performed using the `irtQ::est_irt()` function,
     the parameters of pretest items are estimated while the parameters
     of operational items are fixed.

     To implement FIPC, you must provide the following arguments to `irtQ::est_irt()`:
     - `x`: The item metadata, including both operational and pretest items.
     - `data`: The examinee response data corresponding to the item metadata.
     - `fipc = TRUE`: Enables fixed item parameter calibration.
     - `fipc.method`: Specifies the FIPC method to be used (e.g., `"MEM"`).
     - `fix.loc`: A vector indicating the positions of the operational items to be fixed.

     Optionally, you may estimate the empirical histogram and scale of the latent
     ability distribution by setting `EmpHist = TRUE`. If `EmpHist = FALSE`,
     a normal prior is assumed and its scale is updated iteratively during the EM algorithm.

     For additional details on implementing FIPC, refer to the documentation
     for `irtQ::est_irt()`.

## 3. Pretest Item Calibration with the Fixed Ability Parameter Calibration (FAPC) Method (e.g., Stocking, 1988)

In computerized adaptive testing (CAT), the fixed ability parameter calibration (FAPC) method—also known as Stocking’s Method A (Stocking, 1988)—is one of the simplest and most straightforward approaches for calibrating pretest items. It involves estimating item parameters using maximum likelihood estimation, conditional on known or estimated proficiency values.

FAPC is primarily used to place the parameter estimates of pretest items onto the same scale as the operational item parameters. It can also be used to recalibrate operational items when evaluating potential item parameter drift (Chen & Wang, 2016; Stocking, 1988). This method is known to produce accurate and unbiased item parameter estimates when items are randomly administered to examinees, rather than adaptively, which is often the case for pretest items (Ban et al., 2001; Chen & Wang, 2016).

In the \pkg{irtQ} package, FAPC can be conducted in two main steps:

1. Prepare a data set containing both the item response data and the corresponding ability (proficiency) estimates.

     To use the `irtQ::est_item()` function, two input data sets are required:
     - Ability estimates: A numeric vector containing examinees' ability
       (or proficiency) estimates.
     - Item response data: A matrix or data frame containing item responses,
       where rows represent examinees and columns represent items. The order of
       examinees in the response data must exactly match the order of the ability
       estimates.

2. Estimate the item parameters using the `irtQ::est_item()` function.

     The `irtQ::est_item()` function estimates pretest item parameters based on
     provided ability estimates. To use this function, you must specify the
     following arguments:
     - `data`: A matrix or data frame containing examinees' item responses.
     - `score`: A numeric vector of examinees' ability (proficiency) estimates.
     - `model`: A character vector specifying the IRT model for each item
        (e.g., `"1PLM"`, `"2PLM"`, `"3PLM"`, `"GRM"`, `"GPCM"`).
     - `cats`: A numeric vector indicating the number of score categories for
        each item. For dichotomous items, use 2.
     - `D`: A scaling constant (typically 1.702) to align the logistic
        function with the normal ogive model.

     For additional details on implementing FAPC, refer to the documentation
     for `irtQ::est_item()`.

## 4. The Process of Evaluating the IRT Model–Data Fit

Evaluating how well an item response theory (IRT) model fits observed response data is a critical step in psychometric analysis. The `irtQ` package provides both statistical and graphical tools for evaluating item-level model fit. These include traditional fit statistics (e.g., X², G², infit, outfit, and S-X²) and diagnostic residual plots.

Model fit evaluation using `irtQ` typically involves the following three steps:

### 1. Prepare a data set for model fit analysis

Before conducting the IRT model fit analysis, three key data sets must be prepared:

1. **Item metadata**: A data frame containing item-level information, including:
   - Item ID
   - Number of score categories
   - IRT model specification
   - Calibrated item parameters  
   
   You can either construct this data frame manually or generate it using the `irtQ::shape_df()` function. Additionally, if item parameters were estimated using other IRT software (e.g., BILOG-MG 3, PARSCALE 4, flexMIRT, or the `mirt` R package), you can import them using the corresponding `irtQ::bring.*()` functions (e.g., `irtQ::bring.flexmirt()`, `irtQ::bring.bilog()`).

2. **Ability estimates**: A numeric vector of examinees’ estimated proficiency values.

3. **Response data**: A matrix or data frame in which rows represent examinees and columns represent items. The order of examinees in this matrix must exactly match that of the ability estimates, and the column order must match the item metadata.

### 2. Compute IRT item fit statistics using `irtQ::irtfit()`

The `irtQ::irtfit()` function calculates widely used item fit statistics, including:

- Chi-square (X²)
- Likelihood-ratio chi-square (G²)
- Infit and outfit statistics

To compute X² and G² statistics, the latent ability scale must be divided into several groups. Two grouping methods are available:

- `"equal.width"`: Divides the scale into intervals of equal length
- `"equal.freq"`: Divides the scale into groups with equal numbers of examinees

You must also specify where the expected probabilities of item responses are calculated within each group. Two options are available:

- `"average"`: Uses the average ability estimate within each group
- `"middle"`: Uses the midpoint of each interval

To implement this step, specify the item metadata (`x`), ability estimates (`score`), and response data (`data`) as arguments in the `irtQ::irtfit()` function. If you want to use more or fewer than ten ability groups, adjust the `n.width` argument accordingly. If the response data contain missing values, specify the missing value code using the `missing` argument.

Upon execution, the function returns item fit statistics and contingency tables used to compute the X² and G² statistics.

Note that the model-fit evaluation using the S-X² statistic can be implemented using the `irtQ::sx2_fit()` function.  

### 3. Draw residual plots using the `plot()` method

After obtaining fit statistics using the `irtQ::irtfit()` function, you can use the `plot()` method to visualize residuals for individual items. Two types of plots are available: raw residual plot and standardized residual plot. 

To generate a plot, specify the item to be examined using the `item.loc` argument. Only one item can be plotted at a time.

The `ci.method` argument controls how confidence intervals are computed in the raw residual plots. Supported methods include:

- `"wald"`: Wald interval based on the normal approximation (Laplace, 1812)
- `"cp"`: Clopper–Pearson interval (Clopper & Pearson, 1934)
- `"wilson"`: Wilson score interval (Wilson, 1927)
- `"wilson.cr"`: Wilson score interval with continuity correction (Newcombe, 1998)

These graphical diagnostics complement the statistical fit measures, allowing for deeper investigation into the adequacy of model fit for individual items.


## 5. Examples of implementing online calibration and evaluating the IRT model-data fit

```{r example, out.width="70%", out.height="50%", message = FALSE, tidy=TRUE, tidy.opts=list(arrow=TRUE, indent=2, width.cutoff=50)}

# Attach the packages
library(irtQ)

##---------------------------------------------------------------------------
## 1. Item parameter estimation for a linear test form
##---------------------------------------------------------------------------

## Step 1: Prepare response data for the reference group
## Import the "-prm.txt" output file from flexMIRT
meta_true <- system.file("extdata", "flexmirt_sample-prm.txt", package = "irtQ")

# Extract item metadata using `irtQ::bring.flexmirt()`
# This will serve as the base test form for later pretest item examples
x_new <- irtQ::bring.flexmirt(file = meta_true, "par")$Group1$full_df

# Extract items 1 to 40 to define the linear test form used in this illustration
x_ref <- x_new[1:40, ]

# Generate true ability values (N = 2,000) from N(0, 1) for the reference group
set.seed(20)
theta_ref <- rnorm(2000, mean = 0, sd = 1)

# Simulate response data for the linear test form
# Scaling factor D = 1 assumes a logistic IRT model
data_ref <- irtQ::simdat(x = x_ref, theta = theta_ref, D = 1)

## Step 2: Estimate item parameters for the linear test form
## using the following arguments:
# data       = data_ref                              # Response data
# D          = 1                                     # Scaling factor
# model      = c(rep("3PLM", 38), rep("GRM", 2))     # Item models
# cats       = c(rep(2, 38), rep(5, 2))              # Score categories per item
# item.id    = paste0("Ref_I", 1:40)                 # Item IDs
# use.gprior = TRUE                                  # Use prior for guessing parameter
# gprior     = list(dist = "beta", params = c(5, 16))# Prior: Beta(5,16) for g
# Quadrature = c(49, 6)                              # 49 quadrature points from -6 to 6
# group.mean = 0
# group.var  = 1                                     # Fixed latent ability: N(0,1)
# EmpHist    = TRUE                                  # Estimate empirical ability distribution
# Etol       = 1e-3                                  # E-step convergence tolerance
# MaxE       = 500                                   # Max EM iterations
mod_ref <- irtQ::est_irt(
  data       = data_ref,                             
  D          = 1,                                     
  model      = c(rep("3PLM", 38), rep("GRM", 2)),    
  cats       = c(rep(2, 38), rep(5, 2)),             
  item.id    = paste0("Ref_I", 1:40),                 
  use.gprior = TRUE,                                  
  gprior     = list(dist = "beta", params = c(5, 16)),
  Quadrature = c(49, 6),                              
  group.mean = 0,
  group.var  = 1,                                    
  EmpHist    = TRUE,                                 
  Etol       = 1e-3,                                 
  MaxE       = 500)                                  

# Summarize estimation results
irtQ::summary(mod_ref)

# Extract item parameter estimates
est_ref <- mod_ref$par.est
print(est_ref)

##------------------------------------------------------------------------------
## 2. Pretest item calibration using Fixed Item Parameter Calibration (FIPC)
##------------------------------------------------------------------------------

## Step 1: Prepare item metadata for both fixed operational items and pretest items
# Define anchor item positions (items to be fixed)
fixed_pos <- c(1:40)

# Specify IDs, models, and categories for 15 pretest items
# Includes 12 3PLM and 3 GRM items (each GRM has 5 categories)
new_ids <- paste0("New_I", 1:15)
new_models <- c(rep("3PLM", 12), rep("GRM", 3))
new_cats <- c(rep(2, 12), rep(5, 3))

# Construct item metadata using `shape_df_fipc()`. See Details of `shape_df_fipc()`
# for more information
# First 40 items are anchor items (fixed); last 15 are pretest (freely estimated)
meta_fipc <- irtQ::shape_df_fipc(x = est_ref, fix.loc = fixed_pos, item.id = new_ids,
                                 cats = new_cats, model = new_models)

## Step 2: Prepare response data for the new test form
# Generate latent abilities for 2,000 new examinees from N(0.5, 1.3²)
set.seed(21)
theta_new <- rnorm(2000, mean = 0.5, sd = 1.3)

# Simulate response data using true item parameters and true abilities
data_new <- irtQ::simdat(x = x_new, theta = theta_new, D = 1)

## Step 3: Calibrate pretest items using FIPC
##  Fit 3PLM to dichotomous and GRM to polytomous items
## Fix first 40 items and freely estimate the remaining 15 pretest items
## using the following arguments:
# x           = meta_fipc                     # Combined item metadata
# data        = data_new                      # Response data
# D           = 1                             # Scaling constant
# use.gprior  = TRUE                          # Use prior for guessing parameter
# gprior      = list(dist = "beta", params = c(5, 16))  # Prior: Beta(5,16) for g
# Quadrature  = c(49, 6)                      # 49 quadrature points from -6 to 6
# EmpHist     = TRUE                          # Estimate empirical ability distribution
# Etol        = 1e-3                          # E-step convergence tolerance
# MaxE        = 500                           # Max EM iterations
# fipc        = TRUE                          # Enable FIPC
# fipc.method = "MEM"                         # Use Multiple EM cycles
# fix.loc     = c(1:40)                       # Anchor item positions to fix
mod_fipc <- irtQ::est_irt(
  x           = meta_fipc,                    
  data        = data_new,                   
  D           = 1,                            
  use.gprior  = TRUE,                          
  gprior      = list(dist = "beta", params = c(5, 16)),  
  Quadrature  = c(49, 6),                      
  EmpHist     = TRUE,                         
  Etol        = 1e-3,                         
  MaxE        = 500,                           
  fipc        = TRUE,                         
  fipc.method = "MEM",                         
  fix.loc     = c(1:40))                       

# Summarize estimation results
irtQ::summary(mod_fipc)

# Extract item parameter estimates
est_new_fipc <- mod_fipc$par.est
print(est_new_fipc)

# Plot estimated empirical distribution of ability
emphist <- irtQ::getirt(mod_fipc, what="weights")
plot(emphist$weight ~ emphist$theta, xlab="Theta", ylab="Density", type = "h")

##------------------------------------------------------------------------------
## 3. Pretest item calibration using Fixed Ability Parameter Calibration (FAPC)
##------------------------------------------------------------------------------

## Step 1: Prepare response data and ability estimates
# In FAPC, ability estimates are assumed known and fixed.
# Estimate abilities for new examinees using the first 40 fixed operational (anchor) items only.
# Pretest items are not used for scoring, as their parameters are not yet calibrated.

# Estimate abilities using ML method via `irtQ::est_score()`
# Based on fixed anchor item parameters and corresponding responses
# using the following arguments:
# x      = est_ref            # Metadata with operational item parameters
# data   = data_new[, 1:40]   # Responses to anchor items
# D      = 1                  # Scaling constant
# method = "ML"               # Scoring method: Maximum Likelihood
# range  = c(-5, 5)           # Scoring bounds
score_ml <- irtQ::est_score(
  x      = est_ref,            
  data   = data_new[, 1:40],   
  D      = 1,                  
  method = "ML",               
  range  = c(-5, 5))           

# Extract estimated abilities
theta_est <- score_ml$est.theta

## Step 2: Calibrate pretest items using FAPC
# Only the 15 pretest items are included in the calibration
# using the following arguments:
# data       = data_new[, 41:55]                      # Responses to pretest items
# score      = theta_est                              # Fixed ability estimates
# D          = 1                                       # Scaling constant
# model      = c(rep("3PLM", 12), rep("GRM", 3))       # Item models
# cats       = c(rep(2, 12), rep(5, 3))                # Score categories
# item.id    = paste0("New_I", 1:15)                   # Item IDs
# use.gprior = TRUE                                    # Use prior for guessing parameter
# gprior     = list(dist = "beta", params = c(5, 16)   # Prior: Beta(5,16) for g
mod_fapc <- irtQ::est_item(
  data       = data_new[, 41:55],                     
  score      = theta_est,                            
  D          = 1,                                      
  model      = c(rep("3PLM", 12), rep("GRM", 3)),      
  cats       = c(rep(2, 12), rep(5, 3)),                
  item.id    = paste0("New_I", 1:15),                  
  use.gprior = TRUE,                                    
  gprior     = list(dist = "beta", params = c(5, 16))   
)

# Summarize estimation results
irtQ::summary(mod_fapc)

# Extract item parameter estimates
est_new_fapc <- mod_fapc$par.est
print(est_new_fapc)

## ----------------------------------------------------------------------------
# 4. IRT model-data fit evaluation using `irtQ::irtfit()`
## ----------------------------------------------------------------------------

## Step 1: Prepare the data set for IRT model fit analysis
## In this example, we use a simulated mixed-format CAT data set.
## Only items with more than 1,000 non-missing responses are evaluated.

# Identify items with more than 1,000 valid responses
over1000 <- which(colSums(simCAT_MX$res.dat, na.rm = TRUE) > 1000)

# (1) Item metadata
x <- simCAT_MX$item.prm[over1000, ]
dim(x)
print(x[1:10, ])

# (2) Examinees' ability estimates
score <- simCAT_MX$score
length(score)
print(score[1:100])

# (3) Response data
data <- simCAT_MX$res.dat[, over1000]
dim(data)
print(data[1:20, 1:6])

## Step 2: Compute IRT model–data fit statistics
# (1) Using the "equal.width" method to form ability groups
fit1 <- irtfit(
  x = x, score = score, data = data, group.method = "equal.width",
  n.width = 11, loc.theta = "average", range.score = c(-4, 4), 
  D = 1, alpha = 0.05, missing = NA, overSR = 2.5
)

# Inspect the structure of the returned object
names(fit1)

# View the first 10 rows of fit statistics
fit1$fit_stat[1:10, ]

# View the contingency table for the first item (dichotomous)
fit1$contingency.fitstat[[1]]

# (2) Using the "equal.freq" method to form ability groups
fit2 <- irtfit(
  x = x, score = score, data = data, group.method = "equal.freq",
  n.width = 11, loc.theta = "average", range.score = c(-4, 4), 
  D = 1, alpha = 0.05, missing = NA
)

# View the first 10 rows of fit statistics
fit2$fit_stat[1:10, ]

# View the contingency table for the fourth item (polytomous)
fit2$contingency.fitstat[[4]]

## Step 3: Draw residual plots for IRT model–data fit diagnostics
# 1. Dichotomous item
# (1) Both raw and standardized residual plots
plot(x = fit1, item.loc = 1, type = "both", ci.method = "wald", 
     ylim.sr.adjust = TRUE)

# (2) Raw residual plot only
plot(x = fit1, item.loc = 1, type = "icc", ci.method = "wald", 
     ylim.sr.adjust = TRUE)

# (3) Standardized residual plot only
plot(x = fit1, item.loc = 113, type = "sr", ci.method = "wald", 
     ylim.sr.adjust = TRUE)

# 2. Polytomous item
# (1) Both raw and standardized residual plots
plot(x = fit1, item.loc = 113, type = "both", ci.method = "wald", 
     ylim.sr.adjust = TRUE)

# (2) Raw residual plot only, with two columns in layout
plot(x = fit1, item.loc = 113, type = "icc", ci.method = "wald", 
     layout.col = 2, ylim.sr.adjust = TRUE)

# (3) Standardized residual plot only, with four columns in layout
plot(x = fit1, item.loc = 113, type = "sr", ci.method = "wald", 
     layout.col = 4, ylim.sr.adjust = TRUE)
```
