---
title: "Ability Estimation"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

After item calibration, `est_score()` estimates examinee latent ability ($\theta$)
from item response data. The **irtQ** package supports two broad categories of
IRT scoring method, which differ in the information they use from the response data.

### IRT Pattern-Based Scoring

Pattern-based scoring uses each examinee's *full item-response pattern* — the
complete vector of correct/incorrect (or polytomous) responses across all items —
to estimate $\theta$. Because it conditions on the individual responses rather than
only their sum, pattern-based scoring can distinguish examinees who have the same
total score but answered different items correctly. In principle, this makes better
use of the available measurement information [@kolen_tong_2010].

Under the assumption of local independence, the likelihood of an observed response
pattern $\mathbf{U} = (u_1, u_2, \ldots, u_n)$ given ability $\theta$ is

$$L(\theta \mid \mathbf{U}) = \prod_{i=1}^{n} P(U_i = u_i \mid \theta),$$

where $P(U_i = u_i \mid \theta)$ is the category characteristic function for item $i$.
The different pattern-based methods use this likelihood in different ways:

| Method | Key | Description |
|--------|-----|-------------|
| Maximum Likelihood (ML) | `"ML"` | Finds the $\theta$ maximising $L(\theta \mid \mathbf{U})$. Unbiased in large samples but undefined for all-correct or all-incorrect response patterns [@kolen_tong_2010]. |
| ML with Fences (MLF) | `"MLF"` | Augments the likelihood with imaginary "fence" items (a lower fence with a fixed correct response and an upper fence with a fixed incorrect response) to resolve the boundary-score problem while avoiding the shrinkage of Bayesian methods [@han2016]. |
| Weighted Likelihood (WL) | `"WL"` | Multiplies the likelihood by a weighting function derived from the square root of test information before maximizing, yielding estimates that are nearly unbiased to order $O(n^{-1})$ — substantially less biased than both ML and Bayesian modal estimation over the full $\theta$ scale [@warm1989]. |
| Maximum A Posteriori (MAP) | `"MAP"` | Returns the mode of the posterior $p(\theta \mid \mathbf{U}) \propto L(\theta \mid \mathbf{U})\,g(\theta)$, where $g(\theta)$ is a normal prior. Handles boundary scores but shrinks estimates toward the prior mean. |
| Expected A Posteriori (EAP) | `"EAP"` | Returns the mean of the posterior distribution via Gaussian quadrature integration over $\theta$. Also handles boundary scores; has the smallest conditional error variance among pattern-based methods but introduces the most shrinkage [@kolen_tong_2010]. |

A key practical limitation of plain ML is that it yields no finite solution when
all responses to the scored items are correct or all are incorrect (perfect and
zero scores). MLF, WL, MAP, and EAP all handle these boundary patterns, each
through a different mechanism [@han2016].

### IRT Summed-Score Scoring

Summed-score methods map each possible *raw total score*
$X_s = \sum_{i=1}^{n} U_i$ to a $\theta$ estimate rather than using the
full response pattern. Because all examinees with the same total score receive
the same $\theta$ estimate, these methods are transparent and easy for test
users to understand [@kolen_tong_2010]. In particular, Kolen and Tong [-@kolen_tong_2010]
note that the statistical difference between summed-score and pattern-score
estimators is typically smaller in practice than the difference between Bayesian
and non-Bayesian estimators — so the choice between summed-score and
pattern-score scoring need not be driven primarily by accuracy concerns.

An important practical consideration when using these methods is the treatment of missing data. While pattern-based scoring methods handle missing responses (`NA`) flexibly by conditioning only on the observed items, summed-score methods (`"EAP.SUM"` and `"INV.TCC"`) automatically treat any missing responses as incorrect (recoded as 0s) to calculate the total raw score.

| Method | Key | Description |
|--------|-----|-------------|
| EAP for Summed Scores | `"EAP.SUM"` | Computes the Bayesian EAP estimate $\hat{\theta}_{sEAP} = E(\theta \mid X_s)$ for each possible summed-score value using the Lord–Wingersky recursive algorithm [@thissen1995; @thissen_orlando_2001]. Returns a score table mapping every feasible raw score to a $\theta$ estimate and its SE. |
| Inverse TCC | `"INV.TCC"` | Solves $\hat{\theta}_{TCF}$ from the test characteristic function (TCF) equation $\tau_s(\theta) = X_s$ numerically. A non-Bayesian estimator that is monotonically related to $X_s$ and does not depend on the prior distribution [@stocking1996; @kolen_tong_2010]. Standard errors are computed using a recursion-based analytical approach [@lim2020]. Linear interpolation (`intpol = TRUE`) handles scores outside the range where the TCC is invertible (e.g., scores below the sum of guessing parameters in 3PLM). |

```{r load-pkg}
library(irtQ)
set.seed(2026)
```

---

## Key `est_score()` Arguments

Before working through the examples, it is helpful to understand the most
important arguments of `est_score()`:

| Argument | Description |
|----------|-------------|
| `x` | An `est_irt` object (from `irtQ::est_irt()`), or a data frame of item metadata. When an `est_irt` object is provided, the response data embedded in it are used automatically. |
| `data` | Response matrix (rows = examinees, columns = items). **Required** when `x` is a metadata data frame rather than an `est_irt` object. |
| `D` | IRT scaling constant. Use `1` for the logistic metric or `1.702` to approximate the normal-ogive metric. |
| `method` | Scoring method: `"ML"`, `"MLF"`, `"WL"`, `"MAP"`, `"EAP"`, `"EAP.SUM"`, or `"INV.TCC"`. |
| `range` | `c(lower, upper)` bounding the ability scale for iterative methods (`"ML"`, `"MLF"`, `"WL"`, `"MAP"`). Default `c(-5, 5)`. |
| `norm.prior` | `c(mean, sd)` of the normal prior distribution used by `"MAP"`, `"EAP"`, and `"EAP.SUM"`. Default `c(0, 1)`. |
| `nquad` | Number of Gaussian quadrature points for numerical integration (used by `"EAP"` and `"EAP.SUM"`). Default `41`. |
| `fence.a` | Discrimination parameter of the virtual fence items added in `"MLF"`. Default `3`. |
| `fence.b` | Location of the fence items on the $\theta$ scale in `"MLF"`. Defaults to the `range` bounds when `NULL`. |
| `tol` | Convergence tolerance for iterative methods. Default `1e-4`. |
| `max.iter` | Maximum Newton–Raphson iterations. Default `100`. |
| `se` | Logical; compute standard errors? Always returned for `"EAP.SUM"` and `"INV.TCC"`. Default `TRUE`. |
| `intpol` | Logical; apply linear interpolation in `"INV.TCC"` for extreme scores? Default `TRUE`. |
| `range.tcc` | Ability range used for the `"INV.TCC"` interpolation grid. Default `c(-7, 7)`. |

**Return values** differ by method:

- `"ML"`, `"MLF"`, `"WL"`, `"MAP"`, `"EAP"`: a two-column data frame with columns `est.theta` and `se.theta` (one row per examinee).
- `"EAP.SUM"`, `"INV.TCC"`: a list with two elements — `$est.par` (examinee-level estimates including observed sum scores) and `$score.table` (the complete raw-score-to-$\theta$ mapping table).

---

## Setup: Shared Data for All Examples

### Mixed-Format Test (15 Dichotomous + 10 Polytomous Items)

We define a 25-item test (15 dichotomous 3PLM items + 10 polytomous GRM items
with 4 categories) and simulate responses from 800 examinees. This dataset is
used in the **Part 1** and **Part 2** examples for the mixed-format test.

```{r setup-mixed}
# --- Item metadata: 15 3PLM + 10 GRM items ---
meta_mixed <- shape_df(
  par.drm = list(
    a = c(1.0, 1.2, 0.8, 1.4, 0.9, 1.1, 1.3, 0.7, 1.0, 1.2,
          0.9, 1.1, 1.4, 0.85, 1.0),
    b = c(-2.0, -1.5, -1.0, -0.6, -0.2, 0.0, 0.4, 0.8, 1.1, 1.5,
          -1.3, -0.4,  0.5, 1.0,  1.8),
    g = rep(0.15, 15)
  ),
  par.prm = list(
    a = c(1.5, 1.2, 1.0, 1.3, 0.9, 1.1, 0.8, 1.4, 1.2, 1.0),
    d = list(
      c(-1.5, -0.3,  0.9),
      c(-1.2,  0.0,  1.1),
      c(-0.9,  0.4,  1.4),
      c(-1.1, -0.1,  1.0),
      c(-1.3,  0.3,  1.2),
      c(-0.8,  0.5,  1.5),
      c(-1.0,  0.1,  0.9),
      c(-1.4, -0.2,  1.1),
      c(-0.7,  0.6,  1.3),
      c(-1.2,  0.0,  1.0)
    )
  ),
  item.id = c(paste0("BI", 1:15), paste0("PI", 1:10)),
  cats    = c(rep(2, 15), rep(4, 10)),
  model   = c(rep("3PLM", 15), rep("GRM", 10))
)

# --- Simulate 800 examinees from N(0, 1) ---
theta_mixed <- rnorm(800, mean = 0, sd = 1)
resp_mixed  <- simdat(x = meta_mixed, theta = theta_mixed, D = 1.702)

dim(resp_mixed)   # 800 examinees × 25 items
```

### Dichotomous-Only Test (30 Items, 3PLM)

For examples where a simpler, purely dichotomous test is informative — particularly
for summed-score scoring methods where the score table is easiest to interpret —
we additionally prepare a 30-item test with 3PLM items.

```{r setup-dich}
# --- Item metadata: 30 3PLM items ---
meta_dich <- shape_df(
  par.drm = list(
    a = c(1.0, 1.2, 0.8, 1.4, 0.9, 1.1, 1.3, 0.7, 1.0, 1.2,
          0.9, 1.1, 1.4, 0.85, 1.0, 1.2, 0.8, 1.3, 1.0, 0.9,
          1.1, 1.3, 0.8, 1.0, 1.2, 0.9, 1.4, 1.1, 0.7, 1.0),
    b = c(-2.0, -1.5, -1.0, -0.6, -0.2,  0.0,  0.4,  0.8,  1.1,  1.5,
          -1.3, -0.4,  0.5,  1.0,  1.8, -0.8,  0.2,  0.7, -1.1,  1.3,
          -1.8, -0.9,  0.3,  1.2, -0.5,  0.6, -1.4,  0.1,  0.9, -0.3),
    g = rep(0.15, 30)
  ),
  cats  = rep(2, 30),
  model = rep("3PLM", 30)
)

# --- Simulate 500 examinees from N(0, 1) ---
theta_dich <- rnorm(500, mean = 0, sd = 1)
resp_dich  <- simdat(x = meta_dich, theta = theta_dich, D = 1.702)

dim(resp_dich)   # 500 examinees × 30 items
```

---

## Part 1: IRT Pattern-Based Scoring

This section demonstrates all five pattern-based methods. For each method,
examples are shown for both the **mixed-format test** and the
**dichotomous-only test**.

### ML — Maximum Likelihood

ML estimation finds the $\theta$ that maximizes the log-likelihood of the observed
response pattern. It does not use a prior distribution, making it purely
data-driven. A `range` bound is required to prevent $|\hat{\theta}| \to \infty$
for all-correct or all-incorrect response patterns [@kolen_tong_2010].

```{r score-ml-mixed}
# --- Mixed-format test (3PLM + GRM) ---
score_ml_mixed <- est_score(
  x        = meta_mixed,
  data     = resp_mixed,
  D        = 1.702,
  method   = "ML",
  range    = c(-5, 5),
  tol      = 0.0001,
  max.iter = 100,
  se       = TRUE
)

head(score_ml_mixed)
```

```{r score-ml-dich}
# --- Dichotomous-only test (30 3PLM items) ---
score_ml_dich <- est_score(
  x        = meta_dich,
  data     = resp_dich,
  D        = 1.702,
  method   = "ML",
  range    = c(-5, 5),
  tol      = 0.0001,
  max.iter = 100,
  se       = TRUE
)

head(score_ml_dich)
```

### MLF — ML with Fences (Han, 2016)

MLF adds imaginary "fence" items with fixed responses at both ends of the
$\theta$ scale. This makes the log-likelihood unimodal, eliminating the
boundary-score problem of plain ML while producing estimates that are not
shrunk toward a prior mean — unlike MAP or EAP [@han2016].

```{r score-mlf-mixed}
# --- Mixed-format test ---
score_mlf_mixed <- est_score(
  x        = meta_mixed,
  data     = resp_mixed,
  D        = 1.702,
  method   = "MLF",
  range    = c(-5, 5),
  fence.a  = 3.0,    # discrimination of fence items
  fence.b  = NULL,   # fence locations default to range bounds
  se       = TRUE
)

head(score_mlf_mixed)
```

```{r score-mlf-dich}
# --- Dichotomous-only test ---
score_mlf_dich <- est_score(
  x        = meta_dich,
  data     = resp_dich,
  D        = 1.702,
  method   = "MLF",
  range    = c(-5, 5),
  fence.a  = 3.0,
  fence.b  = NULL,
  se       = TRUE
)

head(score_mlf_dich)
```

### WL — Weighted Likelihood (Warm, 1989)

WL multiplies the likelihood by a weighting function $w(\theta)$ derived from the
square root of test information before maximizing. The resulting estimates are
nearly unbiased to order $O(n^{-1})$ — substantially less biased than both plain
ML ($O(n^{-1})$ bias with a positive correlation with $\theta$) and Bayesian estimators 
(also $O(n^{-1})$ but with negative correlation) — across the entire $\theta$
scale, making WL generally preferable to plain ML when an unbiased non-Bayesian
estimate is desired [@warm1989].

```{r score-wl-mixed}
# --- Mixed-format test ---
score_wl_mixed <- est_score(
  x        = meta_mixed,
  data     = resp_mixed,
  D        = 1.702,
  method   = "WL",
  range    = c(-5, 5),
  se       = TRUE
)

head(score_wl_mixed)
```

```{r score-wl-dich}
# --- Dichotomous-only test ---
score_wl_dich <- est_score(
  x        = meta_dich,
  data     = resp_dich,
  D        = 1.702,
  method   = "WL",
  range    = c(-5, 5),
  se       = TRUE
)

head(score_wl_dich)
```

### MAP — Maximum A Posteriori

MAP incorporates a normal prior $g(\theta)$ and returns the **mode** of the
posterior distribution. Compared with EAP, MAP shrinks estimates less strongly
toward the prior mean, and it can still produce estimates outside the bulk of the
prior when the data are informative enough.

```{r score-map-mixed}
# --- Mixed-format test ---
score_map_mixed <- est_score(
  x          = meta_mixed,
  data       = resp_mixed,
  D          = 1.702,
  method     = "MAP",
  range      = c(-5, 5),
  norm.prior = c(0, 1),
  se         = TRUE
)

head(score_map_mixed)
```

```{r score-map-dich}
# --- Dichotomous-only test ---
score_map_dich <- est_score(
  x          = meta_dich,
  data       = resp_dich,
  D          = 1.702,
  method     = "MAP",
  range      = c(-5, 5),
  norm.prior = c(0, 1),
  se         = TRUE
)

head(score_map_dich)
```

### EAP — Expected A Posteriori

EAP returns the **mean** of the posterior distribution, integrating over a
Gaussian quadrature grid. It has the smallest conditional error variance among
pattern-based estimators, but it introduces the most shrinkage toward the prior
mean and is sensitive to the choice of prior distribution, particularly for
short or less reliable tests [@kolen_tong_2010].

```{r score-eap-mixed}
# --- Mixed-format test ---
score_eap_mixed <- est_score(
  x          = meta_mixed,
  data       = resp_mixed,
  D          = 1.702,
  method     = "EAP",
  norm.prior = c(0, 1),
  nquad      = 41,
  se         = TRUE
)

head(score_eap_mixed)
```

```{r score-eap-dich}
# --- Dichotomous-only test ---
score_eap_dich <- est_score(
  x          = meta_dich,
  data       = resp_dich,
  D          = 1.702,
  method     = "EAP",
  norm.prior = c(0, 1),
  nquad      = 41,
  se         = TRUE
)

head(score_eap_dich)
```

### Comparison: ML vs WL vs MAP vs EAP

The following code computes correlations with the true $\theta$ values and mean
absolute errors for both datasets, summarising the practical differences among
the four pattern-based methods.

```{r score-compare-mixed, fig.width=7, fig.height=5}
# --- Mixed-format test: recovery statistics ---
cors_mixed <- c(
  ML  = cor(score_ml_mixed$est.theta,  theta_mixed),
  WL  = cor(score_wl_mixed$est.theta,  theta_mixed),
  MAP = cor(score_map_mixed$est.theta, theta_mixed),
  EAP = cor(score_eap_mixed$est.theta, theta_mixed)
)
round(cors_mixed, 4)

mae_mixed <- c(
  ML  = mean(abs(score_ml_mixed$est.theta  - theta_mixed)),
  WL  = mean(abs(score_wl_mixed$est.theta  - theta_mixed)),
  MAP = mean(abs(score_map_mixed$est.theta - theta_mixed)),
  EAP = mean(abs(score_eap_mixed$est.theta - theta_mixed))
)
round(mae_mixed, 4)
```

```{r score-compare-dich}
# --- Dichotomous-only test: recovery statistics ---
cors_dich <- c(
  ML  = cor(score_ml_dich$est.theta,  theta_dich),
  WL  = cor(score_wl_dich$est.theta,  theta_dich),
  MAP = cor(score_map_dich$est.theta, theta_dich),
  EAP = cor(score_eap_dich$est.theta, theta_dich)
)
round(cors_dich, 4)

mae_dich <- c(
  ML  = mean(abs(score_ml_dich$est.theta  - theta_dich)),
  WL  = mean(abs(score_wl_dich$est.theta  - theta_dich)),
  MAP = mean(abs(score_map_dich$est.theta - theta_dich)),
  EAP = mean(abs(score_eap_dich$est.theta - theta_dich))
)
round(mae_dich, 4)
```

### Providing an `est_irt` Object vs. Separate Metadata

`est_score()` accepts either an `est_irt` object (which embeds the response data
and item parameters) or item metadata and response data provided separately.
When an `est_irt` object is passed to the `x` argument, `est_score()` automatically extracts both the item parameters and the embedded response data. This streamlines the post-calibration scoring workflow by eliminating the need to explicitly supply a separate `data` matrix.

Both workflows produce identical results:

```{r score-alt-input}
# --- Calibrate the mixed-format test first ---
mod_cal <- est_irt(
  data       = resp_mixed,
  D          = 1.702,
  model      = c(rep("3PLM", 15), rep("GRM", 10)),
  cats       = c(rep(2, 15), rep(4, 10)),
  use.gprior = TRUE,
  gprior     = list(dist = "beta", params = c(4, 16)),
  EmpHist    = FALSE,
  Etol       = 0.01,
  MaxE       = 150,
  se         = FALSE,
  verbose    = FALSE
)

# Score directly from the est_irt object (response data are automatically embedded)
score_from_obj <- est_score(
  x          = mod_cal,     # est_irt object
  method     = "EAP",
  norm.prior = c(0, 1),
  nquad      = 41
)

head(score_from_obj)
```

---

## Part 2: IRT Summed-Score Scoring

Summed-score methods assign the same $\theta$ estimate to all examinees with the
same total raw score. They are computationally efficient, straightforward to
communicate to test users, and — per Kolen and Tong [-@kolen_tong_2010] —
typically give results that are statistically comparable to pattern-based methods
when the key choice is between Bayesian and non-Bayesian estimation rather than
between summed-score and pattern-score approaches.

### EAP.SUM — EAP Based on Summed Scores

`"EAP.SUM"` computes the Bayesian EAP estimate $\hat{\theta}_{sEAP}$ for each
possible summed-score value using the Lord–Wingersky recursive algorithm
[@thissen1995; @thissen_orlando_2001], then maps each examinee's observed sum
score to the corresponding table entry.

#### Mixed-format test

```{r score-eapsum-mixed}
score_eapsum_mixed <- est_score(
  x          = meta_mixed,
  data       = resp_mixed,
  D          = 1.702,
  method     = "EAP.SUM",
  norm.prior = c(0, 1),
  nquad      = 41
)

# Individual-level estimates (observed sum score + ability estimate)
head(score_eapsum_mixed$est.par)

# Score table: every feasible raw score mapped to a theta estimate
score_eapsum_mixed$score.table
```

#### Dichotomous-only test

The score table is easiest to interpret for a purely dichotomous test, because the
possible raw scores are simple number-correct values from 0 to $n$.

```{r score-eapsum-dich}
score_eapsum_dich <- est_score(
  x          = meta_dich,
  data       = resp_dich,
  D          = 1.702,
  method     = "EAP.SUM",
  norm.prior = c(0, 1),
  nquad      = 41
)

head(score_eapsum_dich$est.par)
score_eapsum_dich$score.table
```

### INV.TCC — Inverse Test Characteristic Curve

`"INV.TCC"` solves numerically for the $\hat{\theta}_{TCF}$ that satisfies
$\tau_s(\hat{\theta}_{TCF}) = X_s$, where $\tau_s(\theta)$ is the test
characteristic function [@stocking1996; @kolen_tong_2010]. Because this is a
non-Bayesian estimator, it does not depend on a prior distribution and the
resulting scores are monotonically related to raw scores. Linear interpolation
(`intpol = TRUE`) handles raw scores that fall outside the invertible range of
the TCC (for example, when 3PLM items have nonzero guessing parameters, a raw
score equal to the sum of all guessing parameters $g$ cannot be mapped without
interpolation).

#### Mixed-format test

```{r score-invtcc-mixed}
score_invtcc_mixed <- est_score(
  x         = meta_mixed,
  data      = resp_mixed,
  D         = 1.702,
  method    = "INV.TCC",
  intpol    = TRUE,
  range.tcc = c(-7, 5)
)

head(score_invtcc_mixed$est.par)
score_invtcc_mixed$score.table
```

#### Dichotomous-only test

```{r score-invtcc-dich}
score_invtcc_dich <- est_score(
  x         = meta_dich,
  data      = resp_dich,
  D         = 1.702,
  method    = "INV.TCC",
  intpol    = TRUE,
  range.tcc = c(-7, 5)
)

head(score_invtcc_dich$est.par)
score_invtcc_dich$score.table
```

---

## References
