---
title: "Classification Accuracy and Consistency"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Classification Accuracy and Consistency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

In many testing contexts, test scores are used to assign examinees to performance
categories — for example, pass/fail decisions or placement into proficiency levels.
Two indices quantify the quality of such classification decisions:

- **Classification Accuracy (CA)**: the probability that an examinee is correctly
  classified based on their *true* latent ability.
- **Classification Consistency (CC)**: the probability that the same classification
  would result from two independent administrations of the same test.

Both indices are essential for supporting valid score interpretations in high-stakes
settings. Because data from repeated test administrations are rarely available,
single-administration estimation procedures based on IRT models are commonly used.

**irtQ** implements two IRT-based estimation methods:

| Function    | Method                                              | Cut-score scale          |
|-------------|-----------------------------------------------------|--------------------------|
| `cac_lee()` | @lee2010 — conditional summed-score distribution    | Observed score or theta  |
| `cac_rud()` | @rudner2001; @rudner2005 — normal approximation via test information function (TIF) | Theta scale only     |

Both functions support two estimation approaches:

- **D-method (distribution-based)**: integrates conditional indices over a
  population ability distribution specified by quadrature points and weights.
  This approach is appropriate when the empirical ability distribution (e.g.,
  posterior quadrature weights from IRT calibration) is available.
- **P-method (person-based)**: averages conditional indices over individual
  ability estimates from a sample of examinees.

Both functions return a list with five elements:

- `$confusion`: a confusion matrix (rows = true level, columns = expected level),
- `$marginal`: a data frame of marginal CA and CC indices per performance level
  plus an overall marginal row,
- `$conditional`: a data frame of conditional CA and CC for each theta (or node),
- `$prob.level`: a data frame of the probability of being assigned to each level
  for each theta (or node),
- `$cutscore`: the cut scores used in the analysis (on the observed score scale).

```{r load-pkg}
library(irtQ)
set.seed(2026)
```

---

## Setup: Common Item Metadata and Response Data

We use two test forms throughout this vignette.

### Binary test (20 items, 3PLM)

A 20-item dichotomous test calibrated under the three-parameter logistic model (3PLM).

```{r setup-binary}
meta_bin <- shape_df(
  par.drm = list(
    a = c(1.0, 1.2, 0.8, 1.4, 0.9, 1.1, 1.3, 0.7, 1.0, 1.2,
          0.9, 1.1, 1.4, 0.85, 1.0, 1.2, 0.8, 1.1, 1.3, 0.9),
    b = c(-2.0, -1.5, -1.0, -0.6, -0.2, 0.0, 0.4, 0.8, 1.1, 1.5,
          -1.3, -0.4,  0.5,  1.0,  1.8, -0.8, 0.2, 0.7, -1.1, 1.3),
    g = rep(0.15, 20)
  ),
  item.id = paste0("ITEM", 1:20),
  cats    = 2,
  model   = "3PLM"
)

theta_bin <- rnorm(1000, mean = 0, sd = 1)
resp_bin  <- simdat(x = meta_bin, theta = theta_bin, D = 1.702)
```

```{r calibrate-binary, cache=FALSE}
mod_bin <- est_irt(
  data       = resp_bin,
  D          = 1.702,
  model      = "3PLM",
  cats       = 2,
  item.id    = paste0("ITEM", 1:20),
  use.gprior = TRUE,
  gprior     = list(dist = "beta", params = c(4, 16)),
  EmpHist    = FALSE,
  verbose    = FALSE
)
meta_cal_bin <- mod_bin$par.est

# ML ability estimates (needed for the P-method)
score_bin <- est_score(
  x      = meta_cal_bin,
  data   = resp_bin,
  D      = 1.702,
  method = "ML",
  range  = c(-5, 5),
  se     = TRUE
)
```

### Mixed-format test (15 × 3PLM + 5 × GRM, 4 categories)

A mixed-format test combining 15 dichotomous (3PLM) and 5 polytomous items
(Graded Response Model, GRM; 4 ordered categories). The maximum possible
observed summed score is $15 \times 1 + 5 \times 3 = 30$.

```{r setup-mixed}
meta_mix <- shape_df(
  par.drm = list(
    a = c(1.0, 1.2, 0.9, 1.1, 0.8, 1.3, 1.0, 0.9, 1.1, 1.2,
          0.85, 1.1, 1.3, 0.9, 1.0),
    b = c(-1.5, -1.0, -0.5, -0.2, 0.2, 0.5, 0.8, 1.1, 1.4, -1.2,
          -0.4,  0.1,  0.6,  1.0, -0.8),
    g = rep(0.15, 15)
  ),
  par.prm = list(
    a = c(1.5, 1.2, 1.0, 1.3, 0.9),
    d = list(
      c(-1.5, -0.3,  0.9),
      c(-1.2,  0.0,  1.1),
      c(-0.9,  0.4,  1.4),
      c(-1.1, -0.1,  1.0),
      c(-1.3,  0.3,  1.2)
    )
  ),
  item.id = c(paste0("DRM", 1:15), paste0("GRM", 1:5)),
  cats    = c(rep(2, 15), rep(4, 5)),
  model   = c(rep("3PLM", 15), rep("GRM", 5))
)

theta_mix <- rnorm(1000, mean = 0, sd = 1)
resp_mix  <- simdat(x = meta_mix, theta = theta_mix, D = 1.702)
```

```{r calibrate-mixed, cache=FALSE}
mod_mix <- est_irt(
  data       = resp_mix,
  D          = 1.702,
  model      = c(rep("3PLM", 15), rep("GRM", 5)),
  cats       = c(rep(2, 15), rep(4, 5)),
  item.id    = c(paste0("DRM", 1:15), paste0("GRM", 1:5)),
  use.gprior = TRUE,
  gprior     = list(dist = "beta", params = c(4, 16)),
  EmpHist    = FALSE,
  verbose    = FALSE
)
meta_cal_mix <- mod_mix$par.est

# ML ability estimates
score_mix <- est_score(
  x      = meta_cal_mix,
  data   = resp_mix,
  D      = 1.702,
  method = "ML",
  range  = c(-5, 5),
  se     = TRUE
)
```

### Shared quadrature grid

Both methods use the same quadrature grid for D-method examples.

```{r quad-grid}
quad_nodes   <- seq(-4, 4, by = 0.25)
quad_weights <- gen.weight(dist = "norm", mu = 0, sigma = 1, theta = quad_nodes)
```

---

## Part 1: Lee's Method — `cac_lee()`

### Method overview

@lee2010 proposed a general IRT framework for estimating CA and CC indices from
a single test administration. The method can be applied to tests consisting of
dichotomous items, polytomous items, or a mixture of both.

The key quantity is the **conditional summed-score distribution**, $\Pr(X = x \mid \theta)$,
computed using the Lord–Wingersky recursive algorithm [@lord_wingersky1984; @kolen_brennan2004].
This distribution gives the probability that an examinee with ability $\theta$ obtains
each possible observed summed score $x$.

**Cut scores** partition the range of observed summed scores into $K$ performance
levels. The probability that an examinee at ability $\theta$ is assigned to level $k$
is:
$$
p_\theta(k) = \sum_{x \in \text{level } k} \Pr(X = x \mid \theta).
$$

The **conditional CA index** is defined as the probability that the examinee is
correctly classified, that is, assigned to the level that matches their true performance
level $\eta(\theta)$:
$$
\gamma_\theta = p_\theta(\eta(\theta)).
$$
The true performance level $\eta(\theta)$ is determined by comparing the expected
summed score at $\theta$ to the true cut scores.

The **conditional CC index** is defined as the probability that two independent
test administrations yield the same classification decision:
$$
\phi_\theta = \sum_{k=1}^{K} p_\theta(k)^2.
$$

**Marginal** CA and CC indices are obtained by integrating the conditional indices
over the ability distribution $g(\theta)$:
$$
\gamma = \int_{-\infty}^{\infty} \gamma_\theta \, g(\theta) \, d\theta, \qquad
\phi   = \int_{-\infty}^{\infty} \phi_\theta   \, g(\theta) \, d\theta.
$$

If cut scores are specified on the **theta scale** (i.e., `cut.obs = FALSE`), they
are internally converted to the observed summed-score scale using the Test
Characteristic Curve (TCC): $E(X \mid \theta^*) = \sum_i \sum_j j \, \Pr(U_i = j \mid \theta^*)$.

where:
- $X$ is the total observed summed score across the entire test.
- $\theta^*$ is a specific cut score specified on the latent ability scale.
- $i$ is the index for items ($i = 1, 2, \dots, N$, where $N$ is the total number of items).
- $j$ is the index for the item score categories ($j = 0, 1, \dots, K_i - 1$, where $K_i$ is the number of response categories for item $i$). For dichotomous items, $j \in \{0, 1\}$.
- $U_i$ is the item response random variable for item $i$.
- $\Pr(U_i = j \mid \theta^*)$ is the probability of obtaining a score category $j$ on item $i$ given the ability level $\theta^*$, which is modeled by a specific IRT item response function (e.g., 3PLM or GRM).

### Key arguments in `cac_lee()`

| Argument   | Description |
|------------|-------------|
| `x`        | Item metadata data frame |
| `cutscore` | Numeric vector of cut scores; defines $K + 1$ performance levels |
| `weights`  | Two-column matrix of quadrature nodes and weights for D-method (use `gen.weight()`) |
| `theta`    | Numeric vector of individual ability estimates for P-method |
| `D`        | Scaling constant — must match the value used during calibration |
| `cut.obs`  | `TRUE` (default): cut scores on the observed summed-score scale; `FALSE`: on the theta scale (converted internally via TCC) |

Either `theta` or `weights` must be provided (but not both. If both are supplied, 
`weights` takes priority and the D-method is applied).

---

### Example 1: Binary test with `cac_lee()`

#### (1) D-method: integrate over the population ability distribution

Provide a quadrature grid via `weights`. Use `gen.weight()` to construct weights
from a parametric distribution.

```{r cac-lee-binary-d}
# Two cut scores on the observed summed-score scale (range: 0–20)
# Defines three performance levels: [0,8), [8,14), [14,20]
cutscore_obs <- c(8, 14)

# D-method: integrate over the N(0,1) quadrature grid
cac_l_d <- cac_lee(
  x        = meta_cal_bin,
  cutscore = cutscore_obs,
  weights  = quad_weights,     # quadrature grid (D-method)
  D        = 1.702,
  cut.obs  = TRUE              # cut scores on observed scale (default)
)

cac_l_d
```

The output contains:

- `$confusion`: confusion matrix (rows = true performance level, columns = expected
  level under CA). The diagonal entries correspond to correct classifications.
- `$marginal`: marginal CA and CC per level, plus an overall row labelled `"marginal"`.
- `$cutscore`: the cut scores used (always on the observed summed-score scale).

#### (2) P-method: average over individual ability estimates

Provide individual ML ability estimates via the `theta` argument.

```{r cac-lee-binary-p}
# P-method: average over individual ML ability estimates
cac_l_p <- cac_lee(
  x        = meta_cal_bin,
  cutscore = cutscore_obs,
  theta    = score_bin$est.theta,   # individual estimates
  D        = 1.702,
  cut.obs  = TRUE
)

# Confusion matrix (rows = true level, columns = expected level)
cac_l_p$confusion

# Marginal CA and CC indices
cac_l_p$marginal
```

#### (3) Theta-scale cut scores

When cut scores are naturally expressed on the IRT theta scale (e.g., standard-setting
results), set `cut.obs = FALSE`. The function converts them internally to the
observed summed-score scale using the TCC before computing CA and CC.

```{r cac-lee-binary-theta}
# Cut scores on the theta scale
cutscore_theta <- c(-0.5, 0.8)

cac_l_theta <- cac_lee(
  x        = meta_cal_bin,
  cutscore = cutscore_theta,
  theta    = score_bin$est.theta,
  D        = 1.702,
  cut.obs  = FALSE         # cut scores on the theta scale → converted via TCC
)

# Converted cut scores (now on the observed summed-score scale)
cac_l_theta$cutscore

# Marginal CA and CC indices
cac_l_theta$marginal
```

---

### Example 2: Mixed-format test with `cac_lee()`

For a mixed-format test (dichotomous + polytomous items), the maximum possible
observed summed score is $\sum_j (K_j - 1)$, where $K_j$ is the number of
score categories for item $j$. In our test: 15 binary items (max 15) + 5 four-category
items (max $3 \times 5 = 15$) → max = 30.

```{r cac-lee-mixed}
# Cut scores on the observed summed-score scale (range: 0–30)
cutscore_mix <- c(10, 22)

# D-method
cac_l_mix_d <- cac_lee(
  x        = meta_cal_mix,
  cutscore = cutscore_mix,
  weights  = quad_weights,
  D        = 1.702,
  cut.obs  = TRUE
)
cac_l_mix_d

# P-method
cac_l_mix_p <- cac_lee(
  x        = meta_cal_mix,
  cutscore = cutscore_mix,
  theta    = score_mix$est.theta,
  D        = 1.702,
  cut.obs  = TRUE
)
cac_l_mix_p$marginal
```

---

## Part 2: Rudner's Method — `cac_rud()`

### Method overview

@rudner2001 and @rudner2005 proposed a simpler approach based on the assumption
that IRT ability estimates are normally distributed around the true ability:
$$
\hat{\theta} \mid \theta \sim N\!\left(\theta,\; \mathrm{SE}^2(\theta)\right),
$$
where $\mathrm{SE}(\theta) = 1 / \sqrt{I(\theta)}$ is the standard error of estimation
derived from the TIF.

Under this assumption, the probability that an examinee with true ability $\theta$
obtains an ability estimate falling in the cut-score interval $[c_{k-1}, c_k)$ is:
$$
p_\theta(k) = \Phi\!\left(\frac{c_k - \theta}{\mathrm{SE}(\theta)}\right)
            - \Phi\!\left(\frac{c_{k-1} - \theta}{\mathrm{SE}(\theta)}\right),
$$
where $\Phi(\cdot)$ is the standard normal cumulative distribution function, and
the boundary cut scores are $c_0 = -\infty$ and $c_K = +\infty$.

Conditional CA and CC are then computed in the same way as in Lee's method:
$$
\gamma_\theta = p_\theta(\eta(\theta)), \qquad
\phi_\theta = \sum_{k=1}^{K} p_\theta(k)^2,
$$
and marginal indices are obtained by averaging over the ability distribution.

> **Important**: Unlike `cac_lee()`, cut scores for `cac_rud()` must **always** be
> specified on the **IRT theta scale** (not the observed summed-score scale).

Standard errors can be supplied in two ways:

1. Pass item metadata via `x` — the function computes SE from the TIF internally.
2. Pass a pre-computed SE vector via `se` (same length as `theta` for P-method,
   or same length as the number of quadrature nodes for D-method).

Either `x` or `se` must be provided.

### Key arguments in `cac_rud()`

| Argument   | Description |
|------------|-------------|
| `x`        | Item metadata data frame (optional if `se` is provided; used to compute SE from TIF) |
| `cutscore` | Numeric vector of cut scores on the **theta scale** |
| `theta`    | Numeric vector of individual ability estimates (P-method) |
| `se`       | Numeric vector of standard errors. If `NULL` and `x` is provided, SE is computed from the TIF |
| `weights`  | Two-column matrix of quadrature nodes and weights (D-method) |
| `D`        | Scaling constant |

---

### Example 3: Binary test with `cac_rud()`

#### (1) D-method: SE computed from item metadata

```{r cac-rud-binary-d}
# Cut scores on the theta scale (matching the metric used in cac_lee theta examples)
cutscore_th <- c(-0.5, 0.8)

# D-method: SE computed internally from the TIF using item metadata
cac_r_d <- cac_rud(
  x        = meta_cal_bin,     # item metadata → SE via TIF
  cutscore = cutscore_th,
  weights  = quad_weights,
  D        = 1.702
)

cac_r_d
```

#### (2) P-method: individual ability estimates + SE from item metadata

When individual ability estimates are available, the P-method averages conditional
indices over the sample of examinees.

```{r cac-rud-binary-p}
# P-method: SE computed internally from item metadata
cac_r_p <- cac_rud(
  x        = meta_cal_bin,          # SE computed from TIF
  cutscore = cutscore_th,
  theta    = score_bin$est.theta,   # individual ML estimates
  D        = 1.702
)

# Alternatively, supply ML-based standard errors directly:
cac_r_p2 <- cac_rud(
  cutscore = cutscore_th,
  theta    = score_bin$est.theta,
  se       = score_bin$se.theta     # individual SEs from ML scoring
)

# Confusion matrix
cac_r_p$confusion
cac_r_p2$confusion

# Marginal CA and CC indices
cac_r_p$marginal
cac_r_p2$marginal
```

Note that `cac_r_p` and `cac_r_p2` yield **nearly** identical results because ML-based SEs 
are theoretically equivalent to $1/\sqrt{I(\hat\theta)}$. **However, slight discrepancies 
may occur in practice due to how extreme boundary values are handled.** When using `est_score()` 
with ML estimation, the SEs for extreme ability estimates (e.g., artificially bounded 
at -5 or 5 for all-correct or all-incorrect responses)are internally capped at an arbitrary 
large value (e.g., 99.99999) to prevent computational errors. In contrast, computing 
the SE internally from the test information function (TIF) using `cac_rud()` calculates 
the exact analytical value at those bounded thetas. Therefore, if the sample includes 
examinees with extreme scores, minor differences in the final marginal indices might be observed.

---

### Example 4: Mixed-format test with `cac_rud()`

```{r cac-rud-mixed}
# Cut scores on the theta scale
cutscore_th_mix <- c(-0.5, 0.7)

# D-method
cac_r_mix_d <- cac_rud(
  x        = meta_cal_mix,
  cutscore = cutscore_th_mix,
  weights  = quad_weights,
  D        = 1.702
)
cac_r_mix_d

# P-method
cac_r_mix_p <- cac_rud(
  x        = meta_cal_mix,
  cutscore = cutscore_th_mix,
  theta    = score_mix$est.theta,
  D        = 1.702
)
cac_r_mix_p$marginal
```

---

## Comparing Lee's and Rudner's Methods

```{r method-compare}
# Side-by-side comparison of marginal CA and CC (binary test, P-method)
cat("=== Lee's method (P-method) ===\n")
print(cac_l_p$marginal)

cat("\n=== Rudner's method (P-method, SE from TIF) ===\n")
print(cac_r_p$marginal)
```

The two methods share the same conceptual framework — both estimate CA and CC
by computing, for each ability level, the probabilities of being assigned to each
performance category — but differ in how they model the conditional score distribution
and what metric the cut scores operate on:

| Aspect | Lee (2010) | Rudner (2001, 2005) |
|--------|-----------|---------------------|
| Cut-score metric | Observed summed score (or theta, converted via TCC) | Theta scale only |
| Conditional distribution | Exact conditional summed-score distribution via Lord–Wingersky recursion | Normal approximation: $\hat\theta \mid \theta \sim N(\theta, \text{SE}^2)$ |
| SE source | Implicit (via IRT-based score distribution) | Explicit: $\text{SE}(\theta) = 1/\sqrt{I(\theta)}$ |
| Typical CA/CC values | Generally similar to Rudner's method when IRT fits well | Generally similar to Lee's method when IRT fits well |

**Practical guidance:**

- Use **`cac_lee()`** when cut scores are defined on the observed summed-score scale
  (e.g., raw scores such as 70 out of 100), or when the exact conditional score
  distribution is desired. This method is the more rigorous of the two and is
  applicable to mixed-format assessments with any combination of IRT models.
- Use **`cac_rud()`** when cut scores are expressed on the theta (ability) scale,
  or when standard errors from ability estimation are already available. This method
  is simpler to implement and produces results very similar to Lee's method when
  the normality assumption for ability estimates is reasonable.

---

## References
