---
title: "Classical Test Theory (CTT) Analysis"
output: rmarkdown::html_vignette
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

The rest of **irtQ** centers on item response theory (IRT): item parameters
are estimated on a common ability scale, and item/test statistics generally
depend on those estimated parameters. Many testing programs, however, still
report - or additionally report - classical test theory (CTT) statistics:
simple, total-score-based indices of item difficulty, item discrimination,
and test reliability that require no IRT model fitting at all. irtQ provides
a small set of CTT functions for exactly this purpose, working directly from
raw selected-response data or an already-scored item response matrix. For a
comprehensive treatment of classical test theory, see @crocker_algina1986.

| Function       | Purpose |
|----------------|---------|
| `score_resp()` | Score raw selected-response data (option choices) into a 0/1 matrix (covered in the *Utility Functions* article; briefly revisited below) |
| `freq_score()` | Frequency distribution table (frequency, percentage, cumulative percentage) for a vector of total scores |
| `ctt_distr()`  | Option/category response distribution and point-biserial correlations, including distractor flagging |
| `ctt()`        | Full item- and test-level CTT analysis: difficulty, discrimination, alpha-with-item-removed, reliability, SEM, and the total-score frequency distribution |

```{r load-pkg}
library(irtQ)
```

### From Raw Responses to a Scored Matrix

Both `ctt()` and `freq_score()` expect an already-scored item response matrix
(0/1 for a dichotomous item, or 0, 1, 2, ... for a polytomous item), the same
convention used throughout irtQ (e.g., by `est_irt()`). Raw testing data,
however, usually records the option each examinee actually chose. The
*Utility Functions* article covers `score_resp()` in full - including its
support for numeric, Latin-letter, and general-label (e.g., Korean-syllable)
option-coding schemes - but a short example is worth repeating here, since it
is the natural first step before calling `ctt()` or `freq_score()` on raw
data:

```{r score-resp-bridge}
# Raw responses for 5 examinees on 3 letter-coded items; item 2 has one
# double-marked response ("B,D") and item 3 has one blank (NA) response
raw3 <- data.frame(
  V1 = c("A", "B", NA, "a", "C"),
  V2 = c("D", "D", "D", "B,D", "D"),
  V3 = c("B", "A", "C", "C", NA)
)
key3 <- c("A", "D", "C")

scored3 <- score_resp(data = raw3, key = key3)
scored3$scored          # 0/1 matrix - the input format ctt()/freq_score() expect
scored3$resp_summary     # per-item blank/double-mark/invalid counts
```

`ctt_distr()`, described next, is the one CTT function that instead accepts
raw selected-response data directly (together with an answer key): it calls
`score_resp()` internally so that its distractor analysis and the rest of the
CTT pipeline always agree on how a given response is scored.

The two example data sets used throughout the remainder of this article are
defined once here and reused below:

```{r define-data}
# Data set 1: a realistic, already-scored mixed-format response matrix -
# 1,000 examinees on a 55-item test (50 dichotomous 3PLM items + 5 polytomous
# GRM items), simulated from item parameters imported from a flexMIRT output
# file bundled with irtQ. This mirrors the example used in `?ctt`
flex_sam <- system.file("extdata", "flexmirt_sample-prm.txt", package = "irtQ")
x <- bring.flexmirt(file = flex_sam, "par")$Group1$full_df
set.seed(2026)
theta <- rnorm(1000)
dat_mixed <- simdat(x = x, theta = theta, D = 1)

dim(dat_mixed)   # 1,000 examinees x 55 items
table(x$cats)    # 50 items with cats = 2 (dichotomous), 5 items with cats = 5 (GRM)

# Data set 2: a small dichotomous-only response matrix, used for a simpler,
# easier-to-read walk-through of ctt()'s item- and test-level output
set.seed(1)
dat_bin <- data.frame(matrix(rbinom(300 * 8, 1, 0.6), nrow = 300))
dim(dat_bin)     # 300 examinees x 8 items
```

---

## `freq_score()` — Total-Score Frequency Distribution

`freq_score()` tabulates a vector of total (raw) scores into a frequency
distribution table: frequency, percentage, and cumulative percentage for
every integer score value spanning the observed minimum to maximum (including
any value with zero observations in between, so gaps in the distribution are
visible rather than silently skipped). This is the same table `ctt()` builds
internally for its own `$freq` element, but it is also fully usable on its
own for any vector of integer total scores - for instance, the row sums of
`dat_mixed` computed above:

```{r freq-score-example}
total_mixed <- rowSums(dat_mixed)
freq_out <- freq_score(score = total_mixed)

head(freq_out)
tail(freq_out)
sum(freq_out$freq) == nrow(dat_mixed)   # every examinee accounted for
```

---

## `ctt_distr()` — Option/Category Response Distribution and Distractor Analysis

`ctt_distr()` reports, for every item, how examinees distributed themselves
across response options (or, for already-scored polytomous data, score
categories), together with each option's/category's point-biserial
correlation with the total score. It operates in one of two modes, chosen
automatically by whether a `key` is supplied.

### Selected-response mode: raw option data plus an answer key

When `key` is supplied, `data` is treated as raw, unscored option responses -
exactly the format `score_resp()` expects - and `ctt_distr()` additionally
reports the blank/double-marking rate for each item:

```{r ctt-distr-selected}
raw <- data.frame(
  V1 = c("1", "2", "1", "1", "3", "2", "1", "4"),
  V2 = c("4", "4", "2", "2,4", "4", "1", "4", "4")
)
key <- c(1, 4)
out_distr <- ctt_distr(data = raw, key = key)

out_distr$distr   # one row per item x option: freq, pct, pb_raw, pb_corrected, flag
out_distr$omit    # one row per item: pct_blank, pct_double
```

Each item's option-coding scheme (numeric, Latin-letter, or general label) is
inferred independently from its own `key` value, using exactly the same rule
as `score_resp()` - so a letter-coded or Korean-syllable-coded test form is
analyzed identically to a numerically coded one, once options are mapped.

**A scope note on mixed-format data:** selected-response mode is restricted
to **dichotomous, single-key items only**, since it scores `data` via
`score_resp()` internally, and a selected-response item has exactly one
correct option and therefore cannot yield a partial-credit score by
construction. A mixed dichotomous/polytomous test form cannot be analyzed in
this mode. The scored-category mode described further below has no such
restriction and fully supports a mixed-format test form scored in a single
call - see the *Scored-category mode* section.

### Flagging an attractive distractor

A distractor (a non-key option) whose point-biserial correlation with the
total score is positive - meaning it is chosen disproportionately by
*higher*-scoring examinees rather than lower-scoring ones - is a common
symptom of a miskeyed or ambiguously worded item. `ctt_distr()` flags any
distractor exceeding `crit.distractor` (default `0`):

```{r ctt-distr-flag}
# A single item where the wrong option "2" is chosen only by the 5
# highest-ranked examinees, and the keyed option "1" only by the 5 lowest-
# ranked examinees (total scores supplied externally here purely to make the
# pattern explicit; in practice `total` is usually the test's own total score)
raw_flag <- data.frame(
  target = c("2", "2", "2", "2", "2", "1", "1", "1", "1", "1")
)
total_rank <- c(10, 9, 8, 7, 6, 5, 4, 3, 2, 1)
out_flag <- ctt_distr(data = raw_flag, key = c(1), total = total_rank)

out_flag$distr
```

Option `"2"`'s `pb_raw` is strongly positive, and it is flagged as
"distractor more attractive to high scorers" - exactly the pattern to look
for when reviewing a flagged item for a possible keying error.

### Scored-category mode: already-scored item response data, mixed formats included

When `key` is left at its default (`NULL`), `data` is instead treated as an
already-scored response matrix, and `ctt_distr()` reports each score
category's response proportion and category-total correlation, with no
notion of a "correct" option to flag distractors against (`is_key` is `NA`
and `omit` is `NULL` in this mode).

`cats` is always a *per-item* vector in this mode - there is no assumption
anywhere that every item shares the same number of score categories - so a
single call can score a genuinely **mixed dichotomous/polytomous test form**
together, exactly as `ctt()` does further below. The call below scores all
55 items of `dat_mixed` (the 50 dichotomous items and the 5 polytomous GRM
items) at once, not just a same-format subset:

```{r ctt-distr-scored}
out_cat <- ctt_distr(data = dat_mixed, item.id = x$id, cats = x$cats)

# one dichotomous item (2 category rows) and one polytomous GRM item (5
# category rows), both scored within this same mixed-format call
dichot_id <- x$id[x$cats == 2][1]
poly_id   <- x$id[x$cats > 2][1]
out_cat$distr[out_cat$distr$item %in% c(dichot_id, poly_id), ]
```

Each item's own `cats` value determines how many category rows it
contributes to `out_cat$distr` - two rows for a dichotomous item, five for a
5-category GRM item - regardless of what any other item in the same call
looks like.

For `dichot_id` above, notice that `pb_raw` for category 0 and category 1
are equal in magnitude but opposite in sign - this is expected, not a
computation error. With only two categories, choosing category 0 is the
exact complement of choosing category 1 (every scored response is one or
the other), so the "in category 0" indicator equals one minus the
"in category 1" indicator, and correlating a variable's complement with any
reference variable simply flips the sign of the correlation while leaving
its magnitude unchanged. This exact symmetry is specific to two-category
(dichotomous) items; it does not hold for `poly_id` or any other item with
three or more categories, where no such simple complementary relationship
exists between any two categories.

---

## `ctt()` — Full Item- and Test-Level CTT Analysis

`ctt()` is the main entry point for a complete CTT analysis: for each item it
computes difficulty, an item-total correlation (reported in both raw and
corrected forms), and Cronbach's alpha recomputed with that item removed,
with optional flagging of items whose difficulty or discrimination falls
outside conventional thresholds; at the test level it computes Cronbach's
alpha (raw and standardized), the standard error of measurement (SEM), and
average item difficulty/discrimination; and it tabulates the total-score
frequency distribution via `freq_score()`.

### A simple dichotomous example

```{r ctt-dichotomous}
out_bin <- ctt(data = dat_bin)

out_bin$item    # difficulty, discrimination_raw/corrected, alpha-with-item-removed, flag
out_bin$alpha   # test-level alpha, alpha_std, sem, mean difficulty/discrimination
```

Difficulty for item $j$ is the mean item score divided by its maximum
possible score, $p_j = \bar{X}_j / m_j$ - the familiar proportion-correct
index for a dichotomous item. Discrimination is the Pearson correlation
between the item score and the total score; the **raw** (uncorrected) form
correlates an item with the total score that includes its own contribution,
while the **corrected** (item-excluded) form removes it. Both are always
computed and reported as separate columns; the `correct` argument only
selects which of the two feeds the discrimination-flagging criterion
(`crit.dis`).

### A mixed-format example

`ctt()`'s difficulty definition, $p_j = \bar{X}_j / m_j$, keeps difficulty on
a common 0-1 scale regardless of the number of score categories, so it
generalizes directly to a mixed dichotomous/polytomous test such as
`dat_mixed`:

```{r ctt-mixed}
out_mixed <- ctt(data = dat_mixed, item.id = x$id, cats = x$cats)

# the 5 polytomous (GRM) items, for comparison with the ctt_distr() output above
out_mixed$item[out_mixed$item$cats > 2, ]
```

### Reliability: raw alpha and standardized alpha

Two forms of Cronbach's alpha are always computed and returned in
`ctt()`'s `$alpha` element. Raw alpha uses the standard variance-based
formula
$$\alpha = \frac{k}{k - 1}\left(1 - \frac{\sum_{j} \sigma_j^2}{\sigma_X^2}\right),$$
and reflects the reliability of the actual (unweighted) total score obtained
by simply summing the item scores - the score most tests actually use for
reporting and decisions [@cronbach1951]. Standardized alpha instead first
standardizes every item to unit variance before combining them,
$$\alpha_{std} = \frac{k \bar{r}}{1 + (k - 1)\bar{r}},$$
where $\bar{r}$ is the average pairwise inter-item correlation; it differs
meaningfully from raw alpha mainly when items vary substantially in scale or
format, such as the mix of dichotomous and polytomous items in `dat_mixed`
[@osburn2000]:

```{r ctt-alpha-forms}
out_bin$alpha[, c("alpha", "alpha_std")]     # dichotomous-only: close together
out_mixed$alpha[, c("alpha", "alpha_std")]   # mixed-format: can differ more
```

The standard error of measurement, $SEM = SD(X)\sqrt{1 - \alpha}$ (using raw
alpha), is reported alongside alpha as the standard CTT expression of
measurement precision.

### `print()` and `summary()` methods

`ctt()` returns an object of class `"ctt"`. `print.ctt()` displays a
condensed report; `summary.ctt()` (paired with `print.summary.ctt()`) prints
the full item-level table, the test-level reliability summary, and the
total-score frequency distribution - analogous to how `est_irt()` pairs with
`print.est_irt()`/`summary.est_irt()` elsewhere in irtQ:

```{r ctt-print-summary}
out_bin              # print.ctt(): condensed report
summary(out_bin)     # summary.ctt(): full report
```

---

## Summary

| Function       | Input                                   | Key output                                                   | Used by |
|----------------|------------------------------------------|----------------------------------------------------------------|---------|
| `score_resp()` | Raw option choices + answer key         | 0/1 scored response matrix                                      | `ctt()`, `ctt_distr()` (selected-response mode), `est_irt()`, `est_score()` |
| `freq_score()` | Vector of total (raw) scores            | Frequency / percentage / cumulative-percentage table            | `ctt()` |
| `ctt_distr()`  | Raw option data + key, or a scored matrix | Option/category response distribution, point-biserial correlations, distractor flags | Item and distractor review |
| `ctt()`        | Scored item response matrix             | Item-level statistics, test-level reliability (alpha/SEM), total-score frequency distribution | `print.ctt()`, `summary.ctt()` |

---

## References
