---
title: "DIF Detection"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{DIF Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

Differential item functioning (DIF) occurs when examinees from different groups
who share the same underlying ability have systematically different probabilities
of responding correctly to an item. Detecting DIF is an essential step in
ensuring the fairness and validity of a test.

**irtQ** provides three functions for DIF detection:

| Function   | Method                              | Item Types         | Groups |
|------------|-------------------------------------|--------------------|--------|
| `rdif()`   | Residual-based DIF (RDIF)           | Dichotomous (+ polytomous$^*$) | 2      |
| `grdif()`  | Generalized RDIF (GRDIF)            | Dichotomous (+ polytomous$^*$) | ≥ 2   |
| `catsib()` | CATSIB (modified SIBTEST for CAT)   | Dichotomous only   | 2      |

$^*$ `rdif()` and `grdif()` can accept polytomous item response data, but their
statistical performance for polytomous items has not yet been formally validated.
Research extending and evaluating the RDIF and GRDIF frameworks to polytomous
items is currently ongoing.

All three functions require pooled item response data and group membership
labels. Pooled item parameter estimates and ability estimates are also typically
needed; however, for `catsib()`, item parameters (`x`) are not strictly required
as long as ability estimates and their standard errors are supplied externally.
The `x` argument in `catsib()` is only necessary when ability estimates need to
be computed internally (i.e., `score = NULL`) or when purification is applied,
since ability re-estimation at each purification iteration requires item
parameters.

```{r load-pkg}
library(irtQ)
set.seed(2026)
```

---

## Setup: Simulating DIF Data

We simulate a 40-item dichotomous test (3PLM) with 1,000 examinees per group.
DIF is introduced into six items:

- **Items 1–2**: Uniform DIF — the focal group has a higher difficulty parameter
  ($b + 0.5$) while the discrimination parameter is unchanged.
- **Items 3–4**: Nonuniform DIF — the focal group has a lower discrimination
  parameter ($a \times 0.5$) while the difficulty parameter is unchanged.
- **Items 5–6**: Mixed DIF — the focal group has both a lower discrimination
  parameter ($a \times 0.5$) and a higher difficulty parameter ($b + 0.5$),
  making these items both less discriminating and harder for the focal group.
- **Items 7–40**: No DIF.

The focal group has a slightly lower mean ability ($\mu = -0.3$) than the
reference group ($\mu = 0$), representing a realistic impact condition.

```{r dif-setup}
J   <- 40     # number of items
N_R <- 1000   # reference group size
N_F <- 1000   # focal group size

# Base item parameters (no DIF) shared by both groups
set.seed(100)
a_base <- round(runif(J, 0.7, 1.7), 2)
b_base <- round(runif(J, -4.0, 4.0), 2)
g_base <- rep(0.15, J)

# Reference group item metadata (no DIF)
meta_ref <- shape_df(
  par.drm = list(a = a_base, b = b_base, g = g_base),
  cats    = 2,
  model   = "3PLM"
)

# Focal group item metadata: DIF injected into items 1–6
a_focal      <- a_base
b_focal      <- b_base
b_focal[1:2] <- b_base[1:2] + 0.5          # uniform DIF: harder for focal
a_focal[3:4] <- a_base[3:4] * 0.5          # nonuniform DIF: less discriminating
a_focal[5:6] <- a_base[5:6] * 0.5          # mixed DIF: less discriminating ...
b_focal[5:6] <- b_base[5:6] + 0.5          # ... and harder for focal

meta_foc <- shape_df(
  par.drm = list(a = a_focal, b = b_focal, g = g_base),
  cats    = 2,
  model   = "3PLM"
)

# Simulate ability parameters: reference N(0,1), focal N(-0.3,1)
theta_R <- rnorm(N_R, mean =  0.0, sd = 1)
theta_F <- rnorm(N_F, mean = -0.3, sd = 1)

# Simulate item responses separately for each group
resp_R <- simdat(x = meta_ref, theta = theta_R, D = 1.702)
resp_F <- simdat(x = meta_foc, theta = theta_F, D = 1.702)

# Pool responses and define group membership vector (0 = reference, 1 = focal)
resp_pool <- rbind(resp_R, resp_F)
group_vec <- c(rep(0, N_R), rep(1, N_F))

cat("Pooled data:", nrow(resp_pool), "examinees x", ncol(resp_pool), "items\n")
```

Next, we calibrate item parameters using the pooled data and estimate ability
scores. Because the RDIF framework requires ability estimates based on pooled
(aggregate) item parameters, both calibration and scoring must be performed on
the combined data regardless of group membership [@lim_etal2022].

```{r dif-calibrate, cache=FALSE}
# Calibrate item parameters from the pooled data
mod_pool <- est_irt(
  data       = resp_pool,
  D          = 1.702,
  model      = "3PLM",
  cats       = 2,
  use.gprior = TRUE,
  gprior     = list(dist = "beta", params = c(4, 16)),
  EmpHist    = TRUE,
  verbose    = FALSE
)

meta_pool <- mod_pool$par.est

# Estimate ability using ML method
# ML is recommended for RDIF analysis as it provides point estimates
# unaffected by prior distribution assumptions [@lim_etal2022]
score_pool <- est_score(
  x      = meta_pool,
  data   = resp_pool,
  D      = 1.702,
  method = "ML",
  range  = c(-5, 5)
)$est.theta
```

---

## Part 1: RDIF — Residual-Based DIF (`rdif()`)

### Statistical Framework

The RDIF framework [@lim_etal2022] detects DIF by comparing item-level
residuals—defined as the difference between observed and model-expected item
scores—between the reference and focal groups. For a dichotomously scored item,
the residual for examinee $h$ is:

$$r_h = x_h - P_h(\hat{\theta}_h)$$

where $x_h \in \{0, 1\}$ is the observed response and $P_h(\hat{\theta}_h)$ is
the IRT model-predicted probability of a correct response. Three statistics are
computed from these residuals:

**RDIF$_R$** — targets **uniform DIF** via differences in mean raw residuals:

$$\text{RDIF}_R = \frac{1}{N_F}\sum_{j=1}^{N_F} r_{Fj} - \frac{1}{N_R}\sum_{i=1}^{N_R} r_{Ri}$$

Under the null hypothesis of no DIF, $\text{RDIF}_R$ asymptotically follows a
normal distribution $\mathcal{N}(\mu_{\text{RDIF}_R},\, \sigma^2_{\text{RDIF}_R})$.

**RDIF$_S$** — targets **nonuniform DIF** via differences in mean squared residuals:

$$\text{RDIF}_S = \frac{1}{N_F}\sum_{j=1}^{N_F} r_{Fj}^2 - \frac{1}{N_R}\sum_{i=1}^{N_R} r_{Ri}^2$$

Under the null hypothesis, $\text{RDIF}_S$ also asymptotically follows a normal
distribution $\mathcal{N}(\mu_{\text{RDIF}_S},\, \sigma^2_{\text{RDIF}_S})$.

**RDIF$_{RS}$** — a joint Wald-type statistic that detects **both uniform and
nonuniform DIF** simultaneously. It is based on the bivariate normality of
$(\text{RDIF}_R, \text{RDIF}_S)$ and under the null hypothesis asymptotically
follows a $\chi^2$ distribution with 2 degrees of freedom.

The analytic expressions for the means and variances of $\text{RDIF}_R$ and
$\text{RDIF}_S$ are derived from the IRT model-predicted probabilities (see
@lim_etal2022, Appendix A for details). Because these moments are computed
analytically rather than empirically, the RDIF framework is highly
computationally efficient.

> **Practical guidance:** $\text{RDIF}_{RS}$ is the recommended primary
> detection criterion because it is sensitive to both types of DIF
> simultaneously [@lim_etal2022]. Use $\text{RDIF}_R$ and $\text{RDIF}_S$
> to characterize the *type* of DIF after a significant $\text{RDIF}_{RS}$
> flag.

### Key arguments of `rdif()`

- `x`: Item metadata data frame containing pooled item parameter estimates.
- `data`: Matrix of pooled item response data (rows = examinees, columns = items).
- `score`: Numeric vector of pooled ability estimates. If `NULL`, `rdif()`
  estimates them internally using the method specified in `method`.
- `group`: Vector of group membership labels (length = number of examinees).
- `focal.name`: The label identifying the focal group in `group`.
- `D`: Scaling constant (use the same value as in calibration).
- `alpha`: Significance level for hypothesis testing (default `0.05`).
- `purify`: Logical; whether to apply iterative purification (default `FALSE`).
- `purify.by`: Statistic used for purification: `"rdifrs"`, `"rdifr"`, or `"rdifs"`.
- `max.iter`: Maximum number of purification iterations (default `10`).
- `method`: Scoring method for ability re-estimation during purification.
  `"ML"` is recommended (default).

### Example 1: RDIF without purification

```{r rdif-nopurify}
rdif_npur <- rdif(
  x          = meta_pool,
  data       = resp_pool,
  score      = score_pool,
  group      = group_vec,
  focal.name = 1,          # 1 = focal group
  D          = 1.702,
  alpha      = 0.05,
  purify     = FALSE,
  verbose    = FALSE
)

# Summary output
print(rdif_npur)
```

```{r rdif-nopurify-detail}
# Full table of RDIF statistics for all items
rdif_npur$no_purify$dif_stat

# Items flagged by each statistic
rdif_npur$no_purify$dif_item
```

The output contains a `dif_stat` data frame with the following columns for
each item: `rdifr` (RDIF$_R$ statistic), `z.rdifr` (standardized RDIF$_R$),
`rdifs` (RDIF$_S$ statistic), `z.rdifs` (standardized RDIF$_S$), `rdifrs`
(RDIF$_{RS}$ statistic), `p.rdifr`, `p.rdifs`, `p.rdifrs` (corresponding
p-values), and group sample sizes. Items with p-values below `alpha` are
flagged as DIF.

### Example 2: RDIF with purification

When DIF items are present in the test, they can contaminate the ability
estimates used to compute the RDIF statistics, inflating Type I error rates
for DIF-free items. Iterative purification addresses this by progressively
removing flagged DIF items from ability re-estimation until the set of flagged
items stabilizes [@lim_etal2022].

```{r rdif-purify}
rdif_pur <- rdif(
  x          = meta_pool,
  data       = resp_pool,
  score      = score_pool,   # initial ability estimates (pre-purification)
  group      = group_vec,
  focal.name = 1,
  D          = 1.702,
  alpha      = 0.05,
  purify     = TRUE,
  purify.by  = "rdifrs",     # use RDIF_RS to drive purification
  max.iter   = 20,
  method     = "ML",         # re-estimate abilities with ML at each iteration
  range      = c(-5, 5),
  verbose    = FALSE
)

# Summary output
print(rdif_pur)

# DIF items identified after purification
rdif_pur$with_purify$dif_item
```

The `with_purify` component contains the final results after purification
converged. The `n.iter` field reports how many iterations were performed, and
`complete` indicates whether convergence was achieved before reaching `max.iter`.

---

## Part 2: GRDIF — Multiple-Group DIF Detection (`grdif()`)

### Statistical Framework

The GRDIF framework [@lim_etal2024] generalizes the two-group RDIF approach to
simultaneously detect DIF across $G \geq 2$ groups. Three statistics—
$\text{GRDIF}_R$, $\text{GRDIF}_S$, and $\text{GRDIF}_{RS}$—are constructed
by applying a contrast matrix $\mathbf{C}$ to the stacked vector of per-group
mean raw residuals (MRR) and mean squared residuals (MSR), exploiting their
asymptotic multivariate normality.

Under the null hypothesis that no DIF exists between any pair of groups, the
three statistics follow asymptotic $\chi^2$ distributions:

$$\text{GRDIF}_R \xrightarrow{d} \chi^2_{G-1}, \quad
  \text{GRDIF}_S \xrightarrow{d} \chi^2_{G-1}, \quad
  \text{GRDIF}_{RS} \xrightarrow{d} \chi^2_{2(G-1)}$$

where $G$ is the total number of groups. For the two-group special case
($G = 2$), the GRDIF statistics reduce exactly to the original RDIF statistics.

When a significant omnibus result is found, optional post-hoc pairwise RDIF
analyses (controlled via `post.hoc = TRUE`) can identify which specific pairs
of groups exhibit DIF.

> **Note on polytomous items:** Like `rdif()`, `grdif()` also accepts
> polytomous item response data. However, formal validation of the GRDIF
> framework for polytomous items has not yet been published; research on this
> extension is currently ongoing.

### Key arguments of `grdif()`

- `focal.name`: A *vector* of labels for all focal groups (e.g.,
  `focal.name = c("G2", "G3")` when there are two focal groups).
- `post.hoc`: Logical; whether to perform post-hoc pairwise RDIF tests on
  flagged items (default `TRUE`). Useful for identifying which group pairs
  drive the omnibus DIF signal.
- Other arguments (`x`, `data`, `score`, `group`, `D`, `alpha`, `purify`,
  `purify.by`, `max.iter`, `method`) follow the same conventions as `rdif()`.

### Example 1: GRDIF without purification

We extend the simulation to a three-group scenario by adding a second focal
group. Items 1–2 exhibit uniform DIF (b-parameters shifted upward) in both
focal groups, with a larger shift for Focal Group 2.

```{r grdif-setup}
N_g <- 700   # examinees per group

# Ability parameters: G1 = reference, G2 and G3 = focal groups (with impact)
theta_G1 <- rnorm(N_g,  0.0, 1)
theta_G2 <- rnorm(N_g, -0.2, 1)
theta_G3 <- rnorm(N_g, -0.4, 1)

# Item parameters per group
# G1: no DIF (base parameters from the Setup section)
# G2: items 1-2 harder for this focal group (uniform DIF, b + 0.5)
# G3: items 1-2 even harder for this focal group (uniform DIF, b + 0.8)
meta_G1 <- meta_ref
meta_G2 <- meta_ref
meta_G3 <- meta_ref
meta_G2$par.2[1:2] <- meta_ref$par.2[1:2] + 0.5
meta_G3$par.2[1:2] <- meta_ref$par.2[1:2] + 0.8

# Simulate responses for each group
resp_G1 <- simdat(x = meta_G1, theta = theta_G1, D = 1.702)
resp_G2 <- simdat(x = meta_G2, theta = theta_G2, D = 1.702)
resp_G3 <- simdat(x = meta_G3, theta = theta_G3, D = 1.702)

# Pool data and define group membership
resp_3g <- rbind(resp_G1, resp_G2, resp_G3)
grp_3g  <- c(rep("G1", N_g), rep("G2", N_g), rep("G3", N_g))

cat("Pooled data:", nrow(resp_3g), "examinees x", ncol(resp_3g), "items\n")
```

```{r grdif-calibrate, cache=FALSE}
# Calibrate using pooled data
mod_3g <- est_irt(
  data       = resp_3g,
  D          = 1.702,
  model      = "3PLM",
  cats       = 2,
  use.gprior = TRUE,
  gprior     = list(dist = "beta", params = c(4, 16)),
  EmpHist    = TRUE,
  verbose    = FALSE
)

meta_3g  <- mod_3g$par.est

# Estimate pooled ability scores using ML
score_3g <- est_score(
  x      = meta_3g,
  data   = resp_3g,
  D      = 1.702,
  method = "ML",
  range  = c(-5, 5)
)$est.theta
```

```{r grdif-nopurify}
grdif_npur <- grdif(
  x          = meta_3g,
  data       = resp_3g,
  score      = score_3g,
  group      = grp_3g,
  focal.name = c("G2", "G3"),   # both G2 and G3 are focal groups
  D          = 1.702,
  alpha      = 0.05,
  purify     = FALSE,
  post.hoc   = TRUE,            # run pairwise follow-up for flagged items
  verbose    = FALSE
)

# Summary output (omnibus test results)
print(grdif_npur)

# Items flagged by each GRDIF statistic
grdif_npur$no_purify$dif_item
```

```{r grdif-posthoc}
# Post-hoc pairwise RDIF results for DIF-flagged items
# (identifies which specific group pairs drive the DIF signal)
grdif_npur$no_purify$post.hoc
```

The `post.hoc` component reports pairwise RDIF$_R$, RDIF$_S$, and RDIF$_{RS}$
statistics and p-values for each pair of groups among the items that were
flagged by the omnibus test. This helps pinpoint whether DIF occurs between
the reference and one focal group, between the reference and all focal groups,
or between focal groups only.

### Example 2: GRDIF with purification

```{r grdif-purify}
grdif_pur <- grdif(
  x          = meta_3g,
  data       = resp_3g,
  score      = score_3g,       # initial ability estimates (pre-purification)
  group      = grp_3g,
  focal.name = c("G2", "G3"),
  D          = 1.702,
  alpha      = 0.05,
  purify     = TRUE,
  purify.by  = "grdifr",       # use GRDIF_R to drive purification
  max.iter   = 20,
  method     = "ML",           # re-estimate abilities with ML at each iteration
  range      = c(-5, 5),
  post.hoc   = TRUE,
  verbose    = FALSE
)

# Summary output
print(grdif_pur)

# DIF items identified after purification
grdif_pur$with_purify$dif_item

# Post-hoc pairwise RDIF results after purification
grdif_pur$with_purify$post.hoc
```

---

## Part 3: CATSIB (`catsib()`)

### Statistical Framework

CATSIB [@nandakumar_roussos2004] is a modified version of SIBTEST
[@shealy_stout1993] adapted for computerized adaptive testing (CAT)
environments. The procedure estimates the DIF effect size $\hat{\beta}$ by
comparing the observed proportions of correct responses between the reference
and focal groups across matched ability bins.

A key feature of CATSIB is its **regression correction**: because impact
(group mean ability differences) induces stochastic ordering of the two groups
on $\hat{\theta}$, naively matching examinees on $\hat{\theta}$ inflates Type I
error. To address this, CATSIB first transforms each examinee's ability estimate
into a regression-corrected score $\hat{\theta}^*_G$, which estimates the
conditional expectation $E_G[\theta \mid \hat{\theta}]$ separately for each
group $G$ [@shealy_stout1993]:

$$\hat{\theta}^*_G = \bar{\theta}_G + \hat{\rho}^2_G\left(\hat{\theta} - \bar{\hat{\theta}}_G\right)$$

where $\bar{\theta}_G$ is the group mean of $\hat{\theta}$,
$\bar{\hat{\theta}}_G$ is also estimated by $\bar{\theta}_G$, and
$\hat{\rho}^2_G = 1 - \hat{\sigma}^2_{e,G} / \hat{\sigma}^2_{\hat{\theta},G}$
is the estimated reliability in group $G$, with $\hat{\sigma}^2_{e,G}$ being
the mean squared standard error (SE) of ability estimates and
$\hat{\sigma}^2_{\hat{\theta},G}$ being the observed variance of $\hat{\theta}$
in group $G$. Examinees are then matched on $\hat{\theta}^*$ rather than
$\hat{\theta}$.

The corrected scores $\hat{\theta}^*$ are divided into $K$ equal-width ability
bins. Within each bin $k$, the observed proportions of correct responses,
$\hat{P}_{R,k}$ and $\hat{P}_{F,k}$, are computed for the reference and focal
groups, respectively. The DIF effect size is estimated as:

$$\hat{\beta} = \sum_{k=1}^{K} \left[\hat{P}_{R,k} - \hat{P}_{F,k}\right] \hat{p}_k$$

where $\hat{p}_k$ is the weight for bin $k$. By default (`weight.group =
"comb"`), $\hat{p}_k$ is the observed proportion of all examinees (both groups
combined) classified into bin $k$, following the recommendation of
@nandakumar_roussos2004. Alternatively, $\hat{p}_k$ can be defined using only
the focal group distribution (`weight.group = "foc"`) or only the reference
group distribution (`weight.group = "ref"`).

A positive $\hat{\beta}$ indicates that the item is easier for the reference
group than for the focal group (i.e., DIF favoring the reference group). Under
the null hypothesis of no DIF ($H_0: \beta = 0$), the standardized
$\hat{\beta}/\widehat{SE}(\hat{\beta})$ asymptotically follows a standard normal
distribution.

> **Note:** CATSIB requires standard errors (SE) of ability estimates in
> addition to point estimates, because $\hat{\rho}^2_G$ in the regression
> correction uses the measurement error variance. Supply these via the `se`
> argument. When `score` and `se` are provided externally, `x` (item metadata)
> is not required (`x = NULL`), unless purification is applied — in which case
> `x` must be provided for internal ability re-estimation.

### Key arguments of `catsib()`

- `x`: Item metadata data frame. Required when `score = NULL` (ability
  estimated internally) or when `purify = TRUE`. Can be set to `NULL` if
  `score` and `se` are supplied and `purify = FALSE`.
- `se`: Numeric vector of standard errors for the ability estimates (required
  for the regression correction). Obtain via `est_score(..., se = TRUE)$se.theta`.
- `n.bin`: A two-element vector `c(max_bins, min_bins)` controlling the range
  for the number of ability-scale intervals (default `c(80, 10)`).
- `min.binsize`: Minimum number of examinees required in each bin for both
  groups; bins failing this criterion are excluded from the computation
  (default `3`).
- `max.del`: Maximum allowable proportion of examinees excluded during the
  binning process (default `0.075`).
- `weight.group`: Target ability distribution used to compute $\hat{p}_k$:
  `"comb"` (combined reference + focal; default), `"foc"` (focal only), or
  `"ref"` (reference only).

### Example 1: CATSIB without purification

We use the same 40-item dichotomous data simulated in the Setup section.
Since ability estimates and their SEs are supplied externally and no
purification is applied, `x = NULL` is used.

```{r catsib-nopurify}
# Estimate ability with SE (required for the regression correction)
score_se <- est_score(
  x      = meta_pool,
  data   = resp_pool,
  D      = 1.702,
  method = "ML",
  range  = c(-5, 5),
  se     = TRUE
)

# x = NULL: item metadata not required when score/SE are provided externally
# and purify = FALSE
catsib_npur <- catsib(
  x            = NULL,
  data         = resp_pool,
  score        = score_se$est.theta,
  se           = score_se$se.theta,   # SE required for regression correction
  group        = group_vec,
  focal.name   = 1,
  weight.group = "comb",
  D            = 1.702,
  alpha        = 0.05,
  purify       = FALSE,
  verbose      = FALSE
)

# Summary output
print(catsib_npur)

# Items flagged as DIF
catsib_npur$no_purify$dif_item
```

The `dif_stat` data frame contains the estimated $\hat{\beta}$ statistic, its
standard error (`se`), the standardized $\hat{\beta}$ (`z.beta`), and the
corresponding p-value for each item. Items with $|\hat{\beta}|$ significantly
different from zero (at the `alpha` level) are flagged as DIF.

### Example 2: CATSIB with purification

When purification is applied, `catsib()` re-estimates ability parameters
internally at each iteration using the purified item set. This requires `x`
to be provided. Although Nandakumar and Roussos (2004) did not originally
propose a purification procedure for CATSIB, `catsib()` implements one
following the iterative scheme of @lim_etal2022.

```{r catsib-purify}
# x must be provided when purify = TRUE (needed for internal re-scoring)
catsib_pur <- catsib(
  x            = meta_pool,
  data         = resp_pool,
  score        = score_se$est.theta,
  se           = score_se$se.theta,
  group        = group_vec,
  focal.name   = 1,
  weight.group = "comb",
  D            = 1.702,
  alpha        = 0.05,
  purify       = TRUE,
  max.iter     = 20,
  method       = "ML",         # re-estimate abilities with ML at each iteration
  range        = c(-5, 5),
  verbose      = FALSE
)

# Summary output
print(catsib_pur)

# DIF items identified after purification
catsib_pur$with_purify$dif_item
```

### Important Considerations for CATSIB

**Sensitivity to DIF type.** CATSIB, like its predecessor SIBTEST
[@shealy_stout1993], was originally designed and validated for detecting
**uniform DIF** — the condition in which the direction of group differences in
item performance is consistent across all ability levels. The $\hat{\beta}$
statistic accumulates the signed difference $\hat{P}_{R,k} - \hat{P}_{F,k}$
across all ability bins, which means that positive and negative bin-level
differences cancel each other out. As a result, CATSIB has limited sensitivity
to **nonuniform DIF** (where item discrimination differs between groups) or
**mixed DIF** (where both difficulty and discrimination differ), because in
these cases the bin-level differences may partially cancel and yield
$\hat{\beta} \approx 0$ even when DIF is present. If nonuniform or mixed DIF
is suspected, `rdif()` with RDIF$_S$ or RDIF$_{RS}$ is recommended instead.

**Numerical instability of the regression correction under nonuniform or mixed
DIF.** The regression correction (Equation 7 of @nandakumar_roussos2004)
computes, for each group $G$, the reliability estimate:

$$\hat{\rho}^2_G = 1 - \frac{\hat{\sigma}^2_{e,G}}{\hat{\sigma}^2_{\hat{\theta},G}}$$

where $\hat{\sigma}^2_{e,G}$ is the mean squared SE of ability estimates and
$\hat{\sigma}^2_{\hat{\theta},G}$ is the observed variance of $\hat{\theta}$ in
group $G$. When DIF items with substantially reduced discrimination are present, 
their large IRT standard errors may inflate $\hat{\sigma}^2_{e,G}$ for 
the focal group, causing $\hat{\rho}^2_G$ to approach or fall below zero. 
A negative $\hat{\rho}^2_G$ would reverse the direction of the regression 
correction — compressing scores away from the group mean rather than toward 
it — which severely distorts the bin-level matching and leads to inflated Type I error rates.

To prevent this, `catsib()` clamps $\hat{\rho}^2_G$ to the interval $[0.05,\,
1]$:

$$\hat{\rho}^2_G \leftarrow \max\!\left(0.05,\; \min\!\left(1,\; 1 -
\frac{\hat{\sigma}^2_{e,G}}{\hat{\sigma}^2_{\hat{\theta},G}}\right)\right)$$

When the unclamped value falls below 0.05, a warning is issued. This situation
is most likely to occur during purification, when the removal of DIF-flagged
items reduces the number of items used for ability re-estimation, causing
standard errors to grow. If a warning is triggered, results should be
interpreted with caution — or `purify = FALSE` should be considered.

---

## Summary of Function Inputs

All three functions share a common calling pattern:

```r
func(
  x          = <item metadata>,     # pooled item parameters (data frame)
  data       = <response matrix>,   # pooled responses (examinees x items)
  score      = <ability vector>,    # pooled ability estimates
  group      = <group vector>,      # group membership labels
  focal.name = <focal label(s)>,    # which group(s) are focal
  D          = 1.702,               # scaling constant (match calibration)
  alpha      = 0.05,                # significance level
  purify     = TRUE / FALSE,        # apply iterative purification?
  verbose    = FALSE
)
```

Output always contains `$no_purify$dif_stat` (table of statistics for all
items) and `$no_purify$dif_item` (indices of flagged items). When
`purify = TRUE`, corresponding results are also available under `$with_purify`.

---

## References
