---
title: "Getting Started with irtQ"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

## What is irtQ?

**irtQ** is an R package for unidimensional **Item Response Theory (IRT)** analyses.
It supports mixed-format tests containing both dichotomous and polytomous items, and
provides a unified interface for the most common psychometric tasks in educational
and psychological measurement.

### Why IRT? A Brief Motivation

Before using **irtQ**, it helps to understand *why* IRT is preferred over Classical
Test Theory (CTT) for many measurement tasks.

**Classical Test Theory (CTT)** summarizes examinees by their total (or proportion)
correct score and items by their observed difficulty (p-value) and point-biserial
correlation. While straightforward, CTT has well-known limitations:

- **Sample dependence**: item statistics (difficulty, discrimination) change with
  the examinee group, and person scores change with the item set
- **No item-level probability model**: CTT does not tell us the *probability* that
  a specific person answers a specific item correctly
- **Limited equating**: comparing scores across different test forms requires
  strong assumptions

**Item Response Theory (IRT)** addresses these limitations by modeling the probability
of each response as a function of a latent trait $\theta$ (ability, proficiency, or
attitude) and item parameters. Key advantages are:

- **Invariance**: item parameters do not depend on the ability distribution of the
  calibration sample; person parameters do not depend on which particular items were
  administered
- **Item-level modeling**: each item has its own characteristic curve describing
  how response probability changes with $\theta$
- **Common scale**: items and persons are placed on the *same* $\theta$ continuum,
  enabling equating, adaptive testing, and longitudinal measurement

### Core Capabilities of irtQ

**irtQ** covers the full IRT workflow:

**Primary functions:**

- Estimating item parameters — fixed-form calibration, pretest calibration using
  Fixed Item Parameter Calibration (FIPC) and Fixed Ability Parameter Calibration
  (FAPC) approaches, and multiple-group calibration
  (`est_irt()`, `est_mg()`, `est_item()`)
- Estimating examinee abilities — maximum likelihood (ML), weighted likelihood (WLE),
  maximum a posteriori (MAP), expected a posteriori (EAP), EAP summed scoring, and
  inverse test characteristic curve (TCC) scoring (`est_score()`)
- Evaluating item-level model–data fit — $\chi^2$, $G^2$, infit/outfit, $S$-$X^2$
  (`irtfit()`, `sx2_fit()`)
- Detecting differential item functioning (DIF) — residual-based DIF (RDIF),
  generalized residual-based DIF (GRDIF), and simultaneous item bias test 
  modified for CAT (CATSIB) (`rdif()`, `grdif()`, `catsib()`)
- Computing classification accuracy and consistency indices
  (`cac_lee()`, `cac_rud()`)

**Beyond these IRT-based analyses**, irtQ also provides a small set of
classical test theory (CTT) functions — item difficulty, item-total
correlation, Cronbach's alpha, option/distractor response distributions, and
total-score frequency distributions — for scoring and analyzing
selected-response item data without fitting an IRT model
(`ctt()`, `freq_score()`, `ctt_distr()`, `score_resp()`).

**Utilities:**

| Function | Purpose |
|----------|---------|
| `simdat()` | Simulate IRT item response data |
| `info()` | Item and test information functions |
| `traceline()` | Item and test characteristic curves |
| `lwrc()` | Lord–Wingersky recursion (summed-score distributions) |
| `gen.weight()` | Generate quadrature weights from a distribution |
| `covirt()` | IRT-based covariance between items |
| `run_flexmirt()` | Run flexMIRT directly from R |
| `score_resp()` | Score raw selected-response data into a 0/1 matrix |

This vignette introduces the **IRT models** supported by **irtQ** and the **item
metadata** data structure that connects all functions. At the end, a minimal
end-to-end workflow ties everything together.

---

## Installation

Install the released version from CRAN:

```r
install.packages("irtQ")
```

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("hwangQ/irtQ")
```

Load the package:

```{r load-pkg}
library(irtQ)
```

---

## Supported IRT Models

**irtQ** supports the following unidimensional IRT models for dichotomous and
polytomous items.

### Dichotomous Models: 1PLM, 2PLM, and 3PLM

For a binary response $Y \in \{0, 1\}$ from an examinee with latent ability $\theta$,
the probability of a correct response ($Y = 1$) under the three-parameter logistic
model (3PLM) is:

$$P(Y = 1 \mid \theta) = g + \frac{1 - g}{1 + \exp(-D\,a\,(\theta - b))}$$

The four quantities in this equation each have a concrete interpretation:

**$\theta$ — Latent ability (person parameter)**  
The unobserved trait being measured. The $\theta$ scale is typically standardized so
that $\theta \sim N(0, 1)$ in the reference population: $\theta = 0$ is average
ability, $\theta = 1$ is one standard deviation above average, and so on.

**$a$ — Discrimination (item parameter)**  
Controls how steeply the item characteristic curve (ICC) rises. A highly
discriminating item (large $a$) sharply separates examinees just below and just above
the item's difficulty; a weakly discriminating item (small $a$) produces a nearly
flat ICC and provides little information about $\theta$. In practice, $a$ typically
ranges from about 0.5 to 2.5.

**$b$ — Difficulty (item parameter)**  
The point on the $\theta$ scale where the probability of a correct response is
halfway between $g$ and 1, i.e. $P = g + (1-g)/2$. For the 2PLM (where $g = 0$),
this simplifies to the $\theta$ value where $P = 0.5$. Items with $b < 0$ are easy
(most examinees answer correctly); items with $b > 0$ are hard. Values typically
range from about $-3$ to $+3$.

**$g$ — Pseudo-guessing (item parameter)**  
The lower asymptote of the ICC — the probability of a correct response as
$\theta \to -\infty$. For a 5-option multiple-choice item with pure random guessing,
$g = 0.2$. When $g = 0$ the ICC passes through the origin and the model reduces to
the 2PLM or 1PLM.

**$D$ — Scaling constant**  
$D = 1.702$ is a constant that makes the logistic function closely approximate the
normal ogive used in the original IRT formulation. Using $D = 1.702$ is conventional
when working with logistic models; set $D = 1$ to work on the purely logistic scale.
All irtQ functions accept `D` as an argument with default `1.702`.

#### Special cases: 1PLM and 2PLM

| Model | Discrimination | Guessing | Free parameters |
|-------|---------------|----------|-----------------|
| **1PLM** | Fixed to 1 *or* constrained equal across items | $g = 0$ | $b$ only |
| **2PLM** | Freely estimated, varies across items | $g = 0$ | $a$, $b$ |
| **3PLM** | Freely estimated, varies across items | Freely estimated | $a$, $b$, $g$ |

> **Note on 1PLM in irtQ.** When `fix.a.1pl = TRUE` (in `est_irt()`), the
> discrimination is *fixed* to the value `a.val.1pl` (default 1) — this is the
> strict **Rasch model**. When `fix.a.1pl = FALSE` (the default), the common
> discrimination is *estimated* from the data but constrained to be equal across all
> 1PLM items. Both specifications produce only $b$ as the item-differentiating
> parameter, but the second is more flexible when the Rasch assumption of $a = 1$
> does not exactly hold.

---

### Polytomous Models: GRM and GPCM

For items with $K$ **ordered** response categories scored $0, 1, \ldots, K-1$
(e.g., a 4-point rating scale scored 0–3), irtQ supports two models.

#### Graded Response Model (GRM)

The GRM [@samejima1969] models $K - 1$ **cumulative boundary response functions**,
one for each adjacent pair of categories. The cumulative probability of responding
in category $k$ *or higher* is:

$$P^*(Y \ge k \mid \theta) = \frac{1}{1 + \exp(-D\,a\,(\theta - b_k))}, \quad k = 1, \ldots, K-1$$

with the conventions $P^*(Y \ge 0) = 1$ and $P^*(Y \ge K) = 0$.

The probability of responding in exactly category $k$ is then:

$$P(Y = k \mid \theta) = P^*(Y \ge k \mid \theta) - P^*(Y \ge k+1 \mid \theta)$$

**Parameters:**

- $a$: a single discrimination parameter shared across all category boundaries
- $b_1 < b_2 < \cdots < b_{K-1}$: ordered threshold parameters, where $b_k$ is the
  $\theta$ value at which the probability of responding above category $k-1$
  equals 0.5. The ordering constraint ensures that the response probabilities
  are non-negative for all $\theta$.

**Intuition:** Think of the GRM as applying a 2PLM-like curve to each of the $K-1$
adjacent-category boundaries. $b_1$ marks the easy end of the scale (where most
examinees move from category 0 to category 1), and $b_{K-1}$ marks the hard end
(where examinees move into the top category).

**Example (4-category GRM item, $K = 4$):**

- $b_1 = -1.5$: Examinees with $\theta \approx -1.5$ have a 50% chance of scoring
  1 or higher
- $b_2 = 0.0$: Examinees with $\theta \approx 0$ have a 50% chance of scoring
  2 or higher
- $b_3 = 1.2$: Examinees with $\theta \approx 1.2$ have a 50% chance of scoring
  3 (the top category)

---

#### Generalized Partial Credit Model (GPCM)

The GPCM [@muraki1992] generalizes the Partial Credit Model [PCM; @masters1982]
by allowing items to have different discrimination parameters. The probability of
responding in category $k$ is:

$$P(Y = k \mid \theta) =
\frac{\exp\!\left(\displaystyle\sum_{v=0}^{k} D\,a\,(\theta - b_v)\right)}
     {\displaystyle\sum_{h=0}^{K-1} \exp\!\left(\displaystyle\sum_{v=0}^{h} D\,a\,(\theta - b_v)\right)}$$

where the convention $b_0 = 0$ is used so that the numerator for $k = 0$
simplifies to $\exp(0) = 1$.

**Parameters:**

- $a$: a single discrimination parameter for the item
- $b_1, b_2, \ldots, b_{K-1}$: **step parameters** (also called threshold parameters),
  where $b_v$ governs the transition from category $v-1$ to category $v$.
  Unlike GRM thresholds, GPCM step parameters need *not* be ordered.

**Relationship to other models:**

- Setting $a = 1$ for all items gives the **Partial Credit Model (PCM)** [@masters1982]
- The GPCM is therefore a generalization of the PCM that allows item-varying
  discrimination

**How to specify in irtQ:** For an item with $K$ categories, provide a vector of
$K - 1$ step parameters as the `d` argument in `shape_df()`. The convention
$b_0 = 0$ is handled internally; you do not supply $b_0$.

---

### Summary: All Supported Models

| Model | Item type | Parameters | Notes |
|-------|-----------|------------|-------|
| **1PLM** | Dichotomous | $b$ | $a$ fixed or equal-constrained; $g = 0$ |
| **2PLM** | Dichotomous | $a,\, b$ | $g = 0$ |
| **3PLM** | Dichotomous | $a,\, b,\, g$ | All three parameters free |
| **GRM**  | Polytomous  | $a,\, b_1, \ldots, b_{K-1}$ | Thresholds must be ordered |
| **GPCM** | Polytomous  | $a,\, b_1, \ldots, b_{K-1}$ | Step params need not be ordered |
| **PCM**  | Polytomous  | $b_1, \ldots, b_{K-1}$ | GPCM with $a = 1$ |

> **Alias:** `"DRM"` is accepted as a model string for any dichotomous item
> (1PLM, 2PLM, or 3PLM) in contexts where the model is already encoded in the
> metadata.

---

## Item Metadata: The Central Data Structure

Nearly every **irtQ** function accepts a data frame `x` as its first argument.
This **item metadata** frame is a standardized, row-per-item table that encodes
the IRT model and parameter values for each item in a test form. It serves as the
single input format shared across all analyses — calibration, scoring, fit
evaluation, DIF detection, and more — so you only need to build it once.

### Column Structure

| Column | Type | Description |
|--------|------|-------------|
| `id` | character | Unique item identifier |
| `cats` | integer | Number of response categories (2 for dichotomous) |
| `model` | character | IRT model: `"1PLM"`, `"2PLM"`, `"3PLM"`, `"GRM"`, `"GPCM"` |
| `par.1` | numeric | Discrimination parameter ($a$) |
| `par.2` | numeric | Difficulty ($b$) for dichotomous; first threshold ($b_1$) for polytomous |
| `par.3` | numeric | Guessing ($g$) for 3PLM; second threshold ($b_2$) for polytomous; `NA` otherwise |
| `par.4`, ... | numeric | Additional thresholds for polytomous items; `NA` for dichotomous |

The number of `par.*` columns grows with the maximum number of categories across
items. For example, a test containing a GRM item with 5 categories will have columns
`par.1` through `par.5`, with `NA` filling unused slots for dichotomous items.

### A Concrete Example

```{r show-metadata-structure}
# 3 dichotomous items (2PLM) + 2 polytomous items (GRM, 4 categories)
meta_demo <- shape_df(
  par.drm = list(
    a = c(1.2, 0.9, 1.5),
    b = c(-0.5, 0.0, 1.0),
    g = c(NA, NA, NA)         # NA: no guessing for 2PLM
  ),
  par.prm = list(
    a = c(1.8, 1.3),
    d = list(
      c(-1.2, 0.0, 1.2),      # 3 thresholds for item 4 (4 categories)
      c(-0.8, 0.3, 1.5)       # 3 thresholds for item 5 (4 categories)
    )
  ),
  item.id = paste0("I", 1:5),
  cats    = c(2, 2, 2, 4, 4),
  model   = c("2PLM", "2PLM", "2PLM", "GRM", "GRM")
)
meta_demo
```

Reading across the columns:

- Items I1–I3 are 2PLM: `par.1` = $a$, `par.2` = $b$, `par.3` = `NA` (no guessing),
  `par.4` = `NA` (no third threshold)
- Items I4–I5 are GRM with 4 categories: `par.1` = $a$, `par.2`–`par.4` =
  $b_1, b_2, b_3$

---

## Creating Item Metadata with `shape_df()`

`shape_df()` is the primary tool for building item metadata from scratch.
It takes parameter values as named lists and returns a correctly formatted
metadata data frame.

### Example 1: Six dichotomous items (1PLM, 2PLM, 3PLM mixed)

```{r shape-df-ex1}
meta_ex1 <- shape_df(
  par.drm = list(
    a = c(1.0, 1.0, 0.9, 1.2, 0.8, 2.1),     # discrimination
    b = c(0.2, 2.3, -0.4, -2.5, 0.9, 1.2),   # difficulty
    g = c(NA,  NA,   NA,   NA,  0.2, 0.1)     # guessing: NA for 1PLM/2PLM
  ),
  item.id = paste0("ITEM", 1:6),
  cats  = 2,
  model = c("1PLM", "1PLM", "2PLM", "2PLM", "3PLM", "3PLM")
)
meta_ex1
```

**Key points:**

- `par.drm` takes a named list with three vectors: `a` (discrimination), `b`
  (difficulty), and `g` (guessing). Set `g = NA` for 1PLM and 2PLM items.
- Items 1 and 2 are 1PLM: their `a` values are provided in `par.drm` but will be
  handled according to the `fix.a.1pl` argument in estimation functions. The `par.3`
  column in the output is `NA` because there is no guessing for 1PLM items.
- Items 5 and 6 are 3PLM: their guessing parameters are `0.2` and `0.1` respectively.

### Example 2: Mixed-format test (dichotomous + polytomous)

```{r shape-df-ex2}
meta_ex2 <- shape_df(
  par.drm = list(
    a = c(0.8, 2.1),
    b = c(0.9, 1.2),
    g = c(0.2, 0.1)
  ),
  par.prm = list(
    a = c(1.9, 1.3, 0.7),
    d = list(
      c(-1.20, -0.52,  0.25),       # GRM  item: 4 categories → 3 thresholds
      c(-2.20, -1.50, -0.55),       # GPCM item: 4 categories → 3 step params
      c(-0.92, -0.43,  0.49, 1.11)  # GPCM item: 5 categories → 4 step params
    )
  ),
  item.id = paste0("ITEM", 1:5),
  cats  = c(2, 2, 4, 4, 5),
  model = c("3PLM", "3PLM", "GRM", "GPCM", "GPCM")
)
meta_ex2
```

**Key points:**

- `par.prm` takes a named list with `a` (discrimination for polytomous items)
  and `d` (a *list* of threshold/step parameter vectors, one per polytomous item).
- Each vector in `d` must have length $K - 1$ where $K$ is `cats` for that item:
  4-category items need 3 values; 5-category items need 4 values.
- For a **GRM** item, the values in `d` are thresholds $b_1, \ldots, b_{K-1}$,
  which must be in increasing order.
- For a **GPCM** item, the values in `d` are step parameters $b_1, \ldots, b_{K-1}$,
  which need *not* be ordered.
- The `d` vectors for GRM and GPCM items can be mixed within the same `par.prm`
  list — the `model` argument determines which model applies to each item.

### Example 3: Uniform model, auto-assigned IDs

When all items share the same model, a single model string is recycled.
If `item.id` is omitted, IDs are auto-assigned as `V1`, `V2`, …

```{r shape-df-ex3}
meta_ex3 <- shape_df(
  par.drm = list(
    a = c(1.0, 1.0, 0.9, 1.2, 0.8, 2.1),
    b = c(0.2, 2.3, -0.4, -2.5, 0.9, 1.2),
    g = c(0.1, 0.3, 0.1, 0.2, 0.2, 0.1)
  ),
  cats  = 2,
  model = "3PLM"   # recycled for all six items
)
meta_ex3
```

### Example 4: Default parameters (`default.par = TRUE`)

Set `default.par = TRUE` to create a metadata frame with placeholder parameters
without specifying values explicitly. This is useful when you want to calibrate
items from scratch and only need the structure, or as a starting point for
editing. Default values are $a = 1$, $b = 0$, $g = 0.2$ for 3PLM and `NA` for
1PLM/2PLM; polytomous thresholds default to all zeros.

```{r shape-df-ex4}
meta_ex4 <- shape_df(
  cats        = 2,
  model       = c("1PLM", "2PLM", "2PLM", "3PLM", "3PLM"),
  default.par = TRUE
)
meta_ex4
```

### `shape_df()` Argument Summary

| Argument | Description |
|----------|-------------|
| `par.drm` | Named list with `a`, `b`, `g` vectors for dichotomous items |
| `par.prm` | Named list with `a` and `d` (list of threshold vectors) for polytomous items |
| `item.id` | Character vector of item IDs; auto-assigned as `V1`, `V2`, … if omitted |
| `cats` | Integer vector of category counts per item (scalar if all equal) |
| `model` | Character vector of model strings (scalar if all equal) |
| `default.par` | If `TRUE`, fill parameters with built-in defaults (default `FALSE`) |

---

## Importing Item Metadata from External IRT Software

The `bring.*()` family of functions converts output files from external IRT programs
into **irtQ**'s item metadata format. This makes it easy to use parameters calibrated
in another program as the starting point for downstream **irtQ** analyses.

Each function returns a list; the `$full_df` component contains the complete
standardized metadata data frame.

### Importing from flexMIRT

```{r bring-flexmirt}
# Locate the sample flexMIRT -prm.txt file bundled with irtQ
flex_file <- system.file("extdata", "flexmirt_sample-prm.txt", package = "irtQ")

# Import: returns a list keyed by group (e.g., Group1, Group2, ...)
meta_flex <- bring.flexmirt(file = flex_file, type = "par")$Group1$full_df
meta_flex
```

`bring.flexmirt()` returns a **nested list** keyed by group name. Use
`$Group1$full_df`, `$Group2$full_df`, etc. to access each group's metadata.

### Importing from PARSCALE

```{r bring-parscale}
# Locate the sample PARSCALE .PAR file bundled with irtQ
pscale_file <- system.file("extdata", "parscale_sample.PAR", package = "irtQ")

# Import item parameters
meta_pscale <- bring.parscale(file = pscale_file, type = "par")$full_df
meta_pscale
```

### Importing from BILOG-MG

```r
# Import from a BILOG-MG .PAR output file (not run — replace with your path)
meta_bilog <- bring.bilog(file = "output/mytest.PAR", type = "par")$full_df
```

### Importing from the mirt package

```r
# Fit a model with mirt (not run)
# mirt_mod <- mirt::mirt(data = my_data, model = 1, itemtype = "2PL")

# Convert to irtQ metadata
# meta_mirt <- bring.mirt(x = mirt_mod)$full_df
```

### Common Arguments for `bring.*()` Functions

All `bring.*()` functions share two main arguments:

| Argument | Description |
|----------|-------------|
| `file` | Path to the external software output file |
| `type` | `"par"` for item parameter estimates; `"sco"` for ability score estimates |

In practice, replace `system.file(...)` with the actual path to your file:

```r
meta <- bring.flexmirt(file = "output/my_calibration-prm.txt",
                       type = "par")$Group1$full_df
```

---

## Reading Item Metadata from a CSV File

You can also prepare item metadata in a spreadsheet and read it into R as a CSV
file, as long as the column structure matches the **irtQ** specification.

```r
# Read item metadata from a CSV file (not run — replace with your path)
meta_csv <- read.csv("my_item_parameters.csv", stringsAsFactors = FALSE)

# Verify the structure
str(meta_csv)
head(meta_csv)
```

Ensure that:

- Column names are exactly `id`, `cats`, `model`, `par.1`, `par.2`, `par.3`, …
- `id` is character, `cats` is integer, `model` is character, `par.*` are numeric
- Missing threshold or guessing parameters are coded as `NA`
- GRM threshold columns (`par.2`, `par.3`, …) are in increasing order within each row

---

## A Complete End-to-End Workflow

This vignette has focused on **item metadata** — the central data structure that
connects all **irtQ** functions. To show how metadata fits into the broader analysis
pipeline, the example below walks through a minimal end-to-end workflow: define item
metadata → simulate responses → estimate item parameters → estimate abilities →
check recovery. Each step beyond metadata creation is covered in depth in its own
dedicated vignette (see **What's Next?** at the end of this page).

The following example walks through the core **irtQ** workflow from start to finish:
define item metadata → simulate responses → **estimate item parameters** → estimate
abilities → check recovery.

```{r workflow-demo}
set.seed(2026)

# ---- Step 1: Define item metadata (20-item mixed test) ----
# 15 dichotomous 2PLM items + 5 polytomous GRM items (4 categories)
meta_wf <- shape_df(
  par.drm = list(
    a = c(0.8, 1.2, 1.5, 0.9, 1.1, 1.3, 0.7, 1.0, 1.4, 0.85,
          1.1, 0.9, 1.2, 1.0, 1.3),
    b = c(-1.5, -1.0, -0.5, -0.2, 0.0, 0.2, 0.5, 0.8, 1.0, 1.5,
          -1.2, -0.8, -0.3, 0.1, 0.4),
    g = rep(NA, 15)
  ),
  par.prm = list(
    a = c(1.4, 1.1, 0.9, 1.2, 1.0),
    d = list(
      c(-1.5, -0.2,  0.9),
      c(-1.2,  0.1,  1.2),
      c(-0.9,  0.4,  1.5),
      c(-1.1, -0.1,  1.0),
      c(-1.3,  0.3,  1.3)
    )
  ),
  item.id = c(paste0("D", 1:15), paste0("P", 1:5)),
  cats    = c(rep(2, 15), rep(4, 5)),
  model   = c(rep("2PLM", 15), rep("GRM", 5))
)

# ---- Step 2: Simulate response data (500 examinees, N(0,1)) ----
theta_true <- rnorm(500, mean = 0, sd = 1)
resp <- simdat(x = meta_wf, theta = theta_true, D = 1.702)

cat("Response matrix dimensions:", dim(resp), "\n")  # 500 examinees × 20 items
```

```{r workflow-calibrate}
# ---- Step 3: Estimate item parameters with est_irt() ----
mod <- est_irt(
  data    = resp,
  D       = 1.702,
  model   = c(rep("2PLM", 15), rep("GRM", 5)),
  cats    = c(rep(2, 15), rep(4, 5)),
  item.id = c(paste0("D", 1:15), paste0("P", 1:5)),
  EmpHist = FALSE,
  Etol    = 0.001,
  MaxE    = 300,
  se      = TRUE,
  verbose = FALSE
)

# Summarize calibration results
summary(mod)

# Extract estimated item parameters
est_par <- getirt(mod, what = "par.est")
head(est_par)
```

```{r workflow-score}
# ---- Step 4: Estimate examinee abilities with EAP ----
scores <- est_score(
  x      = est_par,      # use calibrated parameters
  data   = resp,
  D      = 1.702,
  method = "EAP"
)

head(scores$est.theta)   # EAP point estimates
head(scores$se.theta)    # posterior standard deviations
```

```{r workflow-recovery, fig.width=7, fig.height=4}
# ---- Step 5: Evaluate ability recovery ----
r <- cor(scores$est.theta, theta_true)

plot(theta_true, scores$est.theta,
     xlab  = expression("True ability (" * theta * ")"),
     ylab  = "EAP estimate",
     main  = paste0("Ability Recovery  (r = ", round(r, 3), ")"),
     pch   = 16, col = rgb(0, 0, 0, 0.25), cex = 0.8)
abline(0, 1, col = "red", lwd = 2, lty = 2)
grid()
```

High correlation ($r > 0.9$) between EAP estimates and the true simulated abilities
confirms that the item calibration and scoring steps are working correctly.

---

## What's Next?

Now that you understand irtQ's item metadata structure and the supported IRT models,
you are ready to explore each analysis in depth:

| Topic | Vignette | Key Functions |
|-------|----------|---------------|
| Item parameter estimation | `vignette("item-parameter-estimation")` | `est_irt()`, `est_mg()`, `est_item()` |
| Ability estimation | `vignette("ability-estimation")` | `est_score()` |
| Model–data fit | `vignette("model-fit-evaluation")` | `irtfit()`, `sx2_fit()` |
| DIF detection | `vignette("dif-detection")` | `rdif()`, `grdif()`, `catsib()` |
| Classification accuracy | `vignette("classification-analysis")` | `cac_lee()`, `cac_rud()` |
| Utilities | `vignette("utilities")` | `simdat()`, `drm()`, `prm()`, `info()`, `traceline()`, `lwrc()`, `gen.weight()`, `covirt()`, `score_resp()` |
| Classical test theory (CTT) analysis | `vignette("ctt-analysis")` | `ctt()`, `freq_score()`, `ctt_distr()` |
| MST panel evaluation | `vignette("mst-panel-evaluation")` | `reval_mst()` |

---

## References
