---
title: "Utility Functions"
output: rmarkdown::html_vignette
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

Beyond the main analysis pipeline, **irtQ** includes a set of utility functions
for simulating item response data, computing information and characteristic
functions, supporting numerical integration over the ability scale, computing
asymptotic standard errors of item parameter estimates, and scoring raw
selected-response data into a 0/1 matrix suitable for downstream analysis.

| Function    | Purpose |
|-------------|---------|
| `simdat()`  | Simulate IRT item response data |
| `drm()`     | Compute correct-response probabilities for dichotomous items (1PLM / 2PLM / 3PLM) |
| `prm()`     | Compute category response probabilities for polytomous items (GRM / GPCM) |
| `info()`    | Item and test information functions (IIF / TIF) |
| `traceline()` | Item and test characteristic curves (ICC / TCC) |
| `lwrc()`    | Lord–Wingersky recursion: conditional summed-score distributions |
| `gen.weight()` | Generate quadrature nodes and weights from a distribution |
| `covirt()`  | Analytical asymptotic variance-covariance matrices of item parameter estimates |
| `score_resp()` | Score raw selected-response data (option choices) into a 0/1 matrix |

```{r load-pkg}
library(irtQ)
set.seed(2026)
```

Throughout this vignette we reuse two item metadata objects defined below.

```{r define-meta}
# 15-item binary test (3PLM)
meta_bin <- shape_df(
  par.drm = list(
    a = c(1.0, 1.2, 0.8, 1.4, 0.9, 1.1, 1.3, 0.7, 1.0, 1.2,
          0.9, 1.1, 1.4, 0.85, 1.0),
    b = c(-2.0, -1.5, -1.0, -0.5, -0.1, 0.2, 0.6, 0.9, 1.2, 1.6,
          -0.8, -0.3,  0.5,  1.0, 1.8),
    g = rep(0.15, 15)
  ),
  item.id = paste0("I", 1:15),
  cats    = 2,
  model   = "3PLM"
)

# Mixed-format test: 5 dichotomous (3PLM) + 3 polytomous (GRM, 4 categories)
meta_mix <- shape_df(
  par.drm = list(
    a = c(1.0, 1.2, 0.9, 1.1, 0.8),
    b = c(-1.0, -0.3, 0.3, 0.8, 1.4),
    g = rep(0.15, 5)
  ),
  par.prm = list(
    a = c(1.5, 1.2, 1.0),
    d = list(
      c(-1.5, -0.2, 1.0),
      c(-1.2,  0.1, 1.3),
      c(-0.9,  0.4, 1.6)
    )
  ),
  item.id = c(paste0("DRM", 1:5), paste0("GRM", 1:3)),
  cats    = c(rep(2, 5), rep(4, 3)),
  model   = c(rep("3PLM", 5), rep("GRM", 3))
)

# Common theta grid used by info(), traceline(), and lwrc()
theta_grid <- seq(-4, 4, by = 0.1)
```

---

## `simdat()` — Simulate Item Response Data

`simdat()` generates item response matrices from known IRT parameters. It is the
primary tool for producing synthetic data for testing and demonstration. It supports
1PLM, 2PLM, and 3PLM for dichotomous items and GRM and GPCM for polytomous
items, including mixed-format tests.

**Two usage modes:**

- **Via item metadata** (`x` argument): pass a `shape_df()` data frame — the
  simplest and recommended approach.
- **Via raw parameter vectors** (`a.drm`, `b.drm`, `g.drm`, `a.prm`, `d.prm`,
  `cats`, `pr.model`): useful when the metadata object has not yet been created,
  or when parameters come directly from external sources.

### Mode 1: Via item metadata (`x` argument)

#### Binary test

```{r simdat-binary}
# 500 examinees drawn from N(0, 1)
theta_bin <- rnorm(500, mean = 0, sd = 1)

# Simulate via item metadata (x argument)
resp_bin <- simdat(x = meta_bin, theta = theta_bin, D = 1.702)

dim(resp_bin)           # rows = examinees, columns = items
head(resp_bin[, 1:8])   # first 8 items, first 6 examinees
colMeans(resp_bin)      # item p-values (proportion correct)
```

#### Mixed-format test

For a mixed-format test, polytomous responses are integers from `0` to
$K_j - 1$, where $K_j$ is the number of score categories for item $j$.

```{r simdat-mixed}
theta_mix <- rnorm(400, mean = 0, sd = 1)
resp_mix  <- simdat(x = meta_mix, theta = theta_mix, D = 1.702)

dim(resp_mix)
head(resp_mix)    # dichotomous (0/1) and polytomous (0–3) responses
```

### Mode 2: Via raw parameter vectors (without `x`)

When a metadata data frame is not available, item parameters can be passed
directly via dedicated arguments. `cats` must always be specified (2 per
dichotomous item, $K_j$ per polytomous item), and `pr.model` must be supplied
for any polytomous items.

#### Binary-only test

For a purely dichotomous test, only `a.drm`, `b.drm`, `cats` (all 2), and
optionally `g.drm` are needed. Omitting `g.drm` corresponds to 1PLM or 2PLM.

```{r simdat-raw-binary}
theta_raw <- rnorm(300, mean = 0, sd = 1)

# 2PLM: no guessing parameter (g.drm omitted)
resp_2pl <- simdat(
  theta = theta_raw,
  a.drm = c(0.9, 1.2, 1.5, 0.8, 1.1),
  b.drm = c(-1.5, -0.5, 0.0, 0.8, 1.5),
  cats  = rep(2, 5),   # all dichotomous
  D     = 1.702
)
dim(resp_2pl)
colMeans(resp_2pl)

# 3PLM: with guessing parameter
resp_3pl <- simdat(
  theta = theta_raw,
  a.drm = c(0.9, 1.2, 1.5, 0.8, 1.1),
  b.drm = c(-1.5, -0.5, 0.0, 0.8, 1.5),
  g.drm = rep(0.15, 5),
  cats  = rep(2, 5),
  D     = 1.702
)
colMeans(resp_3pl)   # p-values are slightly higher due to guessing
```

#### Mixed-format test

When mixing dichotomous and polytomous items, `cats` must encode the correct
category count for **all** items in order, and `pr.model` covers **only** the
polytomous items in the same order.

```{r simdat-raw-mixed}
# Test structure (5 items): DRM1, DRM2, GRM (4 cats), DRM3, GPCM (3 cats)
resp_mix_raw <- simdat(
  theta    = theta_raw,
  a.drm    = c(1.0, 1.2, 0.9),         # slopes for the 3 DRM items
  b.drm    = c(-1.0, 0.0, 1.0),        # difficulties for the 3 DRM items
  g.drm    = rep(0.15, 3),
  a.prm    = c(1.4, 1.1),              # slopes for the 2 polytomous items
  d.prm    = list(
    c(-1.2, -0.1, 1.0),                # 3 threshold params → 4 categories (GRM)
    c(-0.8,  0.6)                      # 2 threshold params → 3 categories (GPCM)
  ),
  cats     = c(2, 2, 4, 2, 3),        # item order: DRM, DRM, GRM, DRM, GPCM
  pr.model = c("GRM", "GPCM"),         # models for polytomous items only
  D        = 1.702
)
dim(resp_mix_raw)
head(resp_mix_raw)   # columns 3 and 5 are polytomous (0–3 and 0–2)
```

### Key `simdat()` arguments

| Argument    | Description |
|-------------|-------------|
| `x`         | Item metadata from `shape_df()` — use this when available |
| `theta`     | Vector of true ability values |
| `D`         | Scaling constant |
| `a.drm`     | Discrimination parameters for dichotomous items (required when `x = NULL`) |
| `b.drm`     | Difficulty parameters for dichotomous items (required when `x = NULL`) |
| `g.drm`     | Guessing parameters; omit for 1PLM/2PLM, use `NA` for mixed 1/2/3PLM tests |
| `a.prm`     | Discrimination parameters for polytomous items |
| `d.prm`     | List of threshold parameter vectors, one element per polytomous item |
| `cats`      | Number of score categories per item (2 for dichotomous); required when `x = NULL` |
| `pr.model`  | IRT model per polytomous item: `"GRM"` or `"GPCM"` |

**Note on the `D` constant:** The scaling constant `D` is used to make the logistic function closely approximate the normal ogive function. It is crucial to use the exact same `D` value (commonly `1.702` or `1.0`) that was utilized during the item parameter calibration phase to avoid any scaling mismatch in the simulated responses or probability computations.

---

## `drm()` and `prm()` — Item Response Probability Functions

`drm()` and `prm()` are the low-level probability engines underlying
`simdat()`, `traceline()`, `info()`, and the internal estimation routines.
They compute item response probabilities directly from IRT parameters and are
useful whenever you need category probabilities at a specific set of theta
values without constructing a full item metadata frame.

### `drm()` — Dichotomous Response Model

`drm()` computes $P(\theta)$, the probability of a correct response, for one
or more dichotomous items under the 1PLM, 2PLM, or 3PLM:
$$
P(\theta) = g + \frac{1 - g}{1 + \exp(-D \cdot a \cdot (\theta - b))},
$$
where $a$ is discrimination, $b$ is difficulty, $g$ is the guessing parameter
(pseudo-chance level), and $D$ is the scaling constant. Setting $g = 0$ gives
the 2PLM; additionally fixing $a$ to be equal across items gives the 1PLM.

The function returns a **matrix** with rows = theta values and columns = items.

```{r drm-basic}
theta_pts <- c(-2, -1, 0, 1, 2)

# 2PLM (g omitted → defaults to 0): two items
P_2pl <- drm(
  theta = theta_pts,
  a     = c(1.0, 1.5),
  b     = c(-0.5, 0.5),
  D     = 1.702
)
P_2pl   # rows = theta, columns = items

# 3PLM: three items with guessing
P_3pl <- drm(
  theta = theta_pts,
  a     = c(0.9, 1.2, 1.5),
  b     = c(-1.0, 0.0, 1.0),
  g     = c(0.10, 0.15, 0.20),
  D     = 1.702
)
round(P_3pl, 4)
```

A useful check: for any 3PLM item, $P(\theta = b) = (1 + g)/2$.

```{r drm-check}
# Verify: P at theta = b equals (1 + g) / 2
P_at_b <- drm(theta = 0.0, a = 1.2, b = 0.0, g = 0.15, D = 1.702)
cat("P(theta=b):", round(as.numeric(P_at_b), 6),
    "  Expected:", (1 + 0.15) / 2, "\n")
```

### `prm()` — Polytomous Response Model

`prm()` computes category response probabilities for a **single** polytomous
item under the GRM or GPCM. It returns a **matrix** with rows = theta values
and columns = score categories (0, 1, ..., $K-1$), and the rows always sum
to 1.

The `d` argument contains $K - 1$ threshold parameters for a $K$-category item:

- **GRM**: `d` holds the $K - 1$ difficulty thresholds $b_k$ at which
  $P(U \geq k \mid \theta) = 0.5$. The categories are ordered so that higher
  theta values yield higher expected scores.
- **GPCM**: `d` holds the $K - 1$ threshold parameters expressed as the item
  location minus the step parameter for each category boundary. The first
  threshold is fixed at 0 internally, so provide $K - 1$ values.

```{r prm-grm}
# GRM: 4-category item (d has K-1 = 3 thresholds)
P_grm <- prm(
  theta    = theta_pts,
  a        = 1.5,
  d        = c(-1.5, -0.2, 1.0),   # 3 thresholds → 4 categories
  D        = 1.702,
  pr.model = "GRM"
)
round(P_grm, 4)          # 5 rows (theta) × 4 columns (score categories)
rowSums(P_grm)           # should all equal 1
```

```{r prm-gpcm}
# GPCM: 4-category item (same parameter count as GRM)
P_gpcm <- prm(
  theta    = theta_pts,
  a        = 1.5,
  d        = c(-1.5, -0.2, 1.0),
  D        = 1.702,
  pr.model = "GPCM"
)
round(P_gpcm, 4)
rowSums(P_gpcm)          # should all equal 1
```

The GRM and GPCM use the same parameter names but produce different probability
profiles. The GRM is appropriate when the response categories are ordered by
increasing difficulty thresholds; the GPCM allows more flexible step structures.

```{r prm-compare}
# Side-by-side at theta = 0
cat("GRM  category probs at theta=0:\n")
round(prm(0, a=1.5, d=c(-1.5,-0.2,1.0), D=1.702, pr.model="GRM"),  4)

cat("GPCM category probs at theta=0:\n")
round(prm(0, a=1.5, d=c(-1.5,-0.2,1.0), D=1.702, pr.model="GPCM"), 4)
```

For the PCM (a special case of GPCM with equal discrimination across items),
set `a = 1`.

### Key `drm()` and `prm()` arguments

| Argument   | `drm()` | `prm()` | Description |
|------------|:-------:|:-------:|-------------|
| `theta`    | ✓ | ✓ | Numeric vector of ability values |
| `a`        | ✓ | ✓ | Discrimination (slope) parameter(s) |
| `b`        | ✓ |   | Difficulty parameter(s) — dichotomous items only |
| `g`        | ✓ |   | Guessing parameter(s); omit for 1PLM/2PLM (defaults to 0) |
| `d`        |   | ✓ | Vector of $K-1$ threshold parameters — polytomous items only |
| `D`        | ✓ | ✓ | Scaling constant (typically `1.702`) |
| `pr.model` |   | ✓ | `"GRM"` or `"GPCM"` |

---

## `info()` — Item and Test Information Functions

`info()` computes the item information function (IIF) and test information
function (TIF) at specified ability values. The TIF is the sum of all IIFs
and is the reciprocal of the squared conditional standard error of estimation:
$\text{TIF}(\theta) = \sum_j \text{IIF}_j(\theta) = 1/\text{SE}^2(\theta)$.

The function returns an object of class `"info"` with three components:

- `$iif`: matrix of item information values (rows = items, columns = theta points),
- `$tif`: numeric vector of TIF values at each theta point,
- `$theta`: the theta grid supplied by the user.

The `plot()` method visualises IIFs, TIF, or the conditional standard error of
estimation (CSEE = $1/\sqrt{\text{TIF}}$).

```{r info-compute}
info_val <- info(
  x     = meta_bin,
  theta = theta_grid,
  D     = 1.702,
  tif   = TRUE       # include TIF in output (default)
)

names(info_val)
head(info_val$tif)               # TIF at first few theta points
info_val$iif[1:5, 1:6]          # IIF: items 1–5, theta points 1–6
```

### Plot: Test Information Function (TIF)

```{r plot-tif}
plot(
  x         = info_val,
  item.loc  = NULL,              # NULL → plot the TIF
  main.text = "Test Information Function (15-item 3PLM)",
  xlab      = expression(theta),
  ylab      = "Information"
)
```

### Plot: Conditional Standard Error of Estimation (CSEE)

```{r plot-csee}
plot(
  x         = info_val,
  item.loc  = NULL,
  csee      = TRUE,              # plot SE(θ) = 1 / √TIF(θ)
  main.text = "Conditional Standard Error of Estimation",
  xlab      = expression(theta),
  ylab      = expression(SE(theta))
)
```

### Plot: IIF for a single item

```{r plot-iif-single}
plot(
  x         = info_val,
  item.loc  = 1,                 # item position (1-indexed)
  main.text = "IIF for Item 1 (3PLM)",
  xlab      = expression(theta),
  ylab      = "Information"
)
```

### Plot: IIFs for multiple items — overlaid and in panels

```{r plot-iif-overlap, fig.width=9, fig.height=4}
# Overlaid in one panel
plot(
  x         = info_val,
  item.loc  = 1:5,
  overlap   = TRUE,
  main.text = "IIFs for Items 1–5 (overlaid)"
)
```

```{r plot-iif-panels, fig.width=10, fig.height=4}
# One panel per item
plot(
  x          = info_val,
  item.loc   = 1:5,
  overlap    = FALSE,
  layout.col = 5,
  main.text  = "IIFs for Items 1–5 (separate panels)"
)
```

### Mixed-format test

`info()` handles polytomous items transparently. The IIF for a polytomous item
reflects the information contributed by the full graded response structure, and
tends to be broader and higher than a comparable dichotomous item because the
multiple ordered categories each contribute information.

```{r info-mixed-tif}
info_mix <- info(x = meta_mix, theta = theta_grid, D = 1.702, tif = TRUE)

# TIF for the mixed-format test
plot(
  x         = info_mix,
  item.loc  = NULL,
  main.text = "TIF: Mixed-Format Test (5 DRM + 3 GRM)"
)
```

#### IIF for a single polytomous item

Items 6–8 in `meta_mix` are the three GRM items. Their IIFs tend to be broader
than those of dichotomous items.

```{r info-mixed-single-poly}
# IIF for GRM item 1 (item 6 in meta_mix)
plot(
  x         = info_mix,
  item.loc  = 6,
  main.text = "IIF for GRM Item 1 (item 6)",
  xlab      = expression(theta),
  ylab      = "Information"
)
```

#### IIFs for multiple items — mixed dichotomous and polytomous

```{r info-mixed-multi-overlap, fig.width=9, fig.height=4}
# Overlaid: compare DRM items vs. GRM items
plot(
  x         = info_mix,
  item.loc  = c(1, 3, 5, 6, 7, 8),   # 3 DRM + 3 GRM
  overlap   = TRUE,
  main.text = "IIFs: DRM Items (1,3,5) vs. GRM Items (6,7,8) — overlaid"
)
```

```{r info-mixed-multi-panels, fig.width=10, fig.height=4}
# Separate panels for the same set of items
plot(
  x          = info_mix,
  item.loc   = c(1, 3, 5, 6, 7, 8),
  overlap    = FALSE,
  layout.col = 3,
  main.text  = "IIFs: DRM Items (1,3,5) and GRM Items (6,7,8)"
)
```

---

## `traceline()` — Item and Test Characteristic Curves

`traceline()` computes the item characteristic curve (ICC), test characteristic
curve (TCC), and category response probabilities for each item over a specified
ability grid.

The function returns an object of class `"traceline"` with four components:

- `$prob.cats`: a named list of data frames, one per item, containing category
  response probabilities (columns `resp.0`, `resp.1`, ...) at each theta value,
- `$icc`: matrix of expected item scores (rows = theta, columns = items),
- `$tcc`: numeric vector of expected total (summed) scores at each theta,
- `$theta`: the theta grid supplied by the user.

For dichotomous items, the ICC equals $P(\theta)$, the probability of a correct
response. For polytomous items, the ICC is the expected item score
$\sum_k k \cdot P(U = k \mid \theta)$.

The `plot()` method is used for visualisation. Two key arguments control what
is drawn:

- `item.loc`: which item(s) to plot; `NULL` plots the TCC for the whole test.
- `score.curve`: `FALSE` (default) draws category response probability curves;
  `TRUE` draws the expected item score curve (a single curve per item).

Note that when `score.curve = FALSE` and `overlap = FALSE`, only a **single**
item can be specified in `item.loc` (one panel per score category). To plot
multiple items simultaneously with `score.curve = FALSE`, set `overlap = TRUE`
(one panel per item, categories overlaid within each panel).

```{r traceline-compute}
trace_bin <- traceline(x = meta_bin, theta = theta_grid, D = 1.702)

names(trace_bin)
head(trace_bin$tcc)               # TCC: expected summed score at each theta
trace_bin$icc[1:6, 1:5]          # ICC: expected item scores, items 1–5
```

### Plot: Test Characteristic Curve (TCC)

```{r plot-tcc}
plot(
  x         = trace_bin,
  item.loc  = NULL,              # NULL → plot TCC
  main.text = "Test Characteristic Curve (15-item 3PLM)",
  xlab      = expression(theta),
  ylab      = "Expected Total Score"
)
```

### Plot: ICCs for a dichotomous item

For a 3PLM item, `score.curve = FALSE` draws two panels — one for each score
category (0 = incorrect, 1 = correct); `score.curve = TRUE` shows only
$P(\theta)$ (the probability of a correct response) as a single curve.

```{r plot-icc-binary-cats, fig.width=6, fig.height=3}
# score.curve = FALSE: separate panels for Score 0 and Score 1
plot(
  x           = trace_bin,
  item.loc    = 2,
  score.curve = FALSE,
  layout.col  = 2,
  main.text   = "ICCs for Item 2 (3PLM) — by category"
)
```

```{r plot-icc-binary-score}
# score.curve = TRUE: single item score curve (= P(correct) for binary items)
plot(
  x           = trace_bin,
  item.loc    = 2,
  score.curve = TRUE,
  main.text   = "Item Score Curve: Item 2 (3PLM)",
  xlab        = expression(theta),
  ylab        = "P(Correct)"
)
```

### Plot: Item score curves for multiple dichotomous items

```{r plot-icc-multi-overlap, fig.width=9, fig.height=4}
# Overlaid in one panel: all five items in a single plot
plot(
  x           = trace_bin,
  item.loc    = 1:5,
  score.curve = TRUE,
  overlap     = TRUE,
  main.text   = "Item Score Curves: Items 1–5 (overlaid)"
)
```

```{r plot-icc-multi-panels, fig.width=10, fig.height=4}
# One panel per item (separate panels)
plot(
  x           = trace_bin,
  item.loc    = 1:5,
  score.curve = TRUE,
  overlap     = FALSE,
  layout.col  = 5,
  main.text   = "Item Score Curves: Items 1–5 (separate panels)"
)
```

### Plot: Category response curves and expected score curve for a polytomous item

For a GRM item, `score.curve = FALSE` draws one curve per score category
(0, 1, 2, 3). Items 6–8 in `meta_mix` are the three GRM items.

```{r traceline-mix-compute}
trace_mix <- traceline(x = meta_mix, theta = theta_grid, D = 1.702)
```

```{r plot-icc-poly-cats, fig.width=8, fig.height=4}
# Category response curves for GRM item 1 (item 6 overall), separate panels
plot(
  x           = trace_mix,
  item.loc    = 6,
  score.curve = FALSE,
  overlap     = FALSE,
  layout.col  = 2,
  main.text   = "Category Response Curves: GRM Item 1 (4 categories)"
)
```

```{r plot-icc-poly-overlap}
# Same curves overlaid in one panel
plot(
  x           = trace_mix,
  item.loc    = 6,
  score.curve = FALSE,
  overlap     = TRUE,
  main.text   = "Category Response Curves (Overlaid): GRM Item 1"
)
```

```{r plot-icc-poly-score}
# Expected item score curve for GRM item 1
plot(
  x           = trace_mix,
  item.loc    = 6,
  score.curve = TRUE,
  main.text   = "Expected Item Score Curve: GRM Item 1",
  ylab.text   = "Expected Score"
)
```

### Plot: Item score curves for multiple items — mixed dichotomous and polytomous

When `score.curve = TRUE`, items of any format can be plotted together because
all ICCs are expressed on the same expected-score metric.

```{r plot-mix-multi-overlap, fig.width=9, fig.height=4}
# Overlaid: DRM items (1–3) and GRM items (6–8) in one panel
plot(
  x           = trace_mix,
  item.loc    = c(1, 2, 3, 6, 7, 8),
  score.curve = TRUE,
  overlap     = TRUE,
  main.text   = "Item Score Curves: DRM (1–3) and GRM (6–8) — overlaid"
)
```

```{r plot-mix-multi-panels, fig.width=10, fig.height=4}
# One panel per item (separate panels)
plot(
  x           = trace_mix,
  item.loc    = c(1, 2, 3, 6, 7, 8),
  score.curve = TRUE,
  overlap     = FALSE,
  layout.col  = 3,
  main.text   = "Item Score Curves: DRM (1–3) and GRM (6–8)"
)
```

### Plot: Category response curves for multiple polytomous items (overlaid per item)

When `score.curve = FALSE` and `overlap = TRUE`, each item gets its own panel
and the category curves within that item are overlaid using different colours.

```{r plot-poly-multi-cats, fig.width=10, fig.height=4}
# One panel per GRM item (6, 7, 8), with all category curves overlaid within each panel
plot(
  x           = trace_mix,
  item.loc    = 6:8,
  score.curve = FALSE,
  overlap     = TRUE,
  layout.col  = 3,
  main.text   = "Category Response Curves: GRM Items 1–3 (overlaid by category)"
)
```

---

## `lwrc()` — Lord–Wingersky Recursion

`lwrc()` implements the Lord–Wingersky recursive algorithm
[@lord_wingersky1984; @kolen_brennan2004], which computes the **conditional
distribution of the observed summed score given $\theta$**:
$$
\Pr(X = s \mid \theta), \quad s = 0, 1, \ldots, \sum_j (K_j - 1).
$$

This distribution is the foundation of IRT-based score equating and is used
internally by `sx2_fit()`, `cac_lee()`, and `est_score()` (EAP summed-score
method). For tests with polytomous items, the algorithm generalizes
straightforwardly to ordered response categories.

The function supports **two input modes**:

- **`x` argument** (item metadata): computes the conditional distribution at
  each specified `theta` value; returns a matrix with rows = summed scores
  ($0, 1, \ldots, J$) and columns = theta levels.
- **`prob` argument** (`x = NULL`): the user provides a category probability
  matrix already evaluated at a particular theta (or a single marginal weight),
  together with a `cats` vector; returns a **named vector** of summed-score
  probabilities. This is useful when category probabilities come from an external
  source or when computing the marginal distribution at a single fixed ability level.

### Mode 1: Using item metadata (`x` argument)

```{r lwrc-binary}
# Conditional summed-score distribution at three ability levels
lwrc_val <- lwrc(
  x     = meta_bin,
  theta = c(-1, 0, 1),
  D     = 1.702
)

dim(lwrc_val)           # (J+1) rows × length(theta) columns: 16 × 3
round(head(lwrc_val), 4)

# Verify: expected summed scores from lwrc() match the TCC from traceline()
score_axis <- 0:(nrow(lwrc_val) - 1)
colSums(lwrc_val * score_axis)
```

```{r lwrc-verify}
# Cross-check with TCC from traceline()
tcc_check <- traceline(x = meta_bin, theta = c(-1, 0, 1), D = 1.702)$tcc
round(tcc_check, 6)
```

The two results agree, confirming that the expected summed score derived from
`lwrc()` matches the TCC value at the same ability level.

### Mode 2: Using a category probability matrix (`x = NULL`)

When `x = NULL`, the user supplies `prob` (rows = items, columns = score
categories) and `cats`. **Importantly, the columns in the `prob` matrix strictly correspond to the ordered score categories (i.e., column 1 is for score 0, column 2 is for score 1, column 3 is for score 2, and so forth).** Empty cells for items with fewer categories should be
filled with `0` or `NA`. This mode returns the **marginal** score distribution
for the single ability level at which the probabilities were computed.

```{r lwrc-prob-binary}
## Example from Kolen & Brennan (2004, p. 183): 3 dichotomous items at a fixed theta
# Each row is one item; columns are P(score=0) and P(score=1)
probs_3items <- matrix(
  c(0.26, 0.74,   # item 1: P(0) = 0.26, P(1) = 0.74
    0.27, 0.73,   # item 2
    0.18, 0.82),  # item 3
  nrow = 3, ncol = 2, byrow = TRUE
)
cats_3items <- c(2, 2, 2)

score_dist <- lwrc(prob = probs_3items, cats = cats_3items)
score_dist       # P(X=0), P(X=1), P(X=2), P(X=3)
sum(score_dist)  # should equal 1
```

```{r lwrc-prob-mixed}
## Mixed-format: 2 dichotomous + 1 four-category + 1 three-category item
p1 <- c(0.3, 0.7, NA,  NA)    # dichotomous: 2 categories (cols: 0, 1)
p2 <- c(0.4, 0.6, NA,  NA)    # dichotomous: 2 categories (cols: 0, 1)
p3 <- c(0.1, 0.3, 0.4, 0.2)   # polytomous:  4 categories (cols: 0, 1, 2, 3)
p4 <- c(0.5, 0.3, 0.2, NA)    # polytomous:  3 categories (cols: 0, 1, 2)

prob_mix <- rbind(p1, p2, p3, p4)
cats_mix  <- c(2, 2, 4, 3)

score_dist_mix <- lwrc(prob = prob_mix, cats = cats_mix)
score_dist_mix   # P(X=0) through P(X=7); max score = 1+1+3+2 = 7
sum(score_dist_mix)
```

---

## `gen.weight()` — Quadrature Weights

`gen.weight()` generates a two-column data frame of quadrature nodes and
normalised weights to be used in numerical integration over the ability scale.
It is used by `est_score()` (EAP), `sx2_fit()`, `cac_lee()`, `cac_rud()`,
and `covirt()`.

Three distribution options are available via the `dist` argument:

- `"norm"`: normal distribution. When `n` is given (no `theta`), Gaussian
  quadrature points and weights are computed via `statmod::gauss.quad.prob()`.
  When `theta` is given, weights are proportional to the normal density
  evaluated at each node.
- `"unif"`: uniform distribution — equally spaced nodes over $[l, u]$ with
  equal weights.
- `"emp"`: empirical distribution — equal weights for user-supplied `theta`
  values.

The returned data frame always has columns named `theta` and `weight`, and
the weights always sum to 1.

```{r gen-weight-norm}
# 41 Gaussian quadrature points from N(0, 1) — most common usage
w_norm <- gen.weight(n = 41, dist = "norm", mu = 0, sigma = 1)
head(w_norm)
sum(w_norm$weight)      # always sums to 1

# Plot the quadrature approximation
plot(w_norm$weight ~ w_norm$theta,
     type = "h", lwd = 2,
     xlab = expression(theta), ylab = "Weight",
     main = "Gaussian Quadrature Approximation of N(0, 1) — 41 nodes")
```

```{r gen-weight-grid}
# Normal weights on a user-defined evenly spaced grid
w_grid <- gen.weight(dist = "norm", mu = 0, sigma = 1,
                     theta = seq(-4, 4, by = 0.25))
nrow(w_grid)   # 33 nodes
sum(w_grid$weight)
```

```{r gen-weight-unif}
# Uniform distribution: 21 equally spaced nodes on [-3, 3]
w_unif <- gen.weight(n = 21, dist = "unif", l = -3, u = 3)
head(w_unif)
```

```{r gen-weight-emp}
# Empirical distribution: equal weights for sample ability estimates
theta_sample <- rnorm(200)
w_emp <- gen.weight(dist = "emp", theta = theta_sample)
nrow(w_emp)              # one node per observation
sum(w_emp$weight)        # sums to 1
```

---

## `covirt()` — Asymptotic Variance-Covariance Matrices of Item Parameters

`covirt()` computes the analytical asymptotic variance-covariance matrices of
item parameter estimates using the formulas developed by @thissen_wainer1982
and extended to polytomous IRT models by @li_lissitz2004. These matrices
provide the asymptotic standard errors (ASEs) of maximum likelihood estimates
**without requiring examinee response data** — only the item parameters and the
sample size used for calibration are needed.

The ASEs obtained analytically represent **lower bounds** of the true standard
errors [@thissen_wainer1982], so they are best used as approximations when
empirical standard errors from calibration software are unavailable.

The function returns a named list with two components:

- `$cov`: a named list of variance-covariance matrices, one per item,
- `$se`: a named list of ASE vectors (square roots of diagonal elements of
  each covariance matrix).

**Notes on PCM items:** The item metadata uses `"GPCM"` to label both the
Generalized Partial Credit Model (GPCM) and the Partial Credit Model (PCM).
PCM items (where discrimination is fixed at 1) require special handling for
variance computation and must be identified via the `pcm.loc` argument.

### Binary test

```{r covirt-binary}
# Compute ASEs for the 15-item 3PLM test, assuming n = 1,000
cov_bin <- covirt(
  x          = meta_bin,
  D          = 1.702,
  nstd       = 1000,
  norm.prior = c(0, 1),
  nquad      = 41
)

# Variance-covariance matrix for item 1
cov_bin$cov[["I1"]]

# ASEs for all items: each vector holds SE(a), SE(b), SE(c)
do.call(rbind, cov_bin$se)
```

### Item-specific sample sizes

When items were calibrated on different samples, pass a vector of sample sizes:

```{r covirt-varying-n}
# Example: items 1–5 calibrated on n=500, items 6–15 on n=2000
n_vec <- c(rep(500, 5), rep(2000, 10))
cov_bin2 <- covirt(x = meta_bin, D = 1.702, nstd = n_vec,
                   norm.prior = c(0, 1), nquad = 41)

# Compare SE(b) for item 1 (n=500) vs item 6 (n=2000)
cat("SE(b) for I1 (n=500):", cov_bin2$se[["I1"]][2], "\n")
cat("SE(b) for I6 (n=2000):", cov_bin2$se[["I6"]][2], "\n")
```

### Mixed-format test

```{r covirt-mixed}
# GRM items in meta_mix: items 6–8 (DRM1–DRM5 come first)
cov_mix <- covirt(
  x          = meta_mix,
  D          = 1.702,
  nstd       = 1500,
  norm.prior = c(0, 1),
  nquad      = 41
)

# ASEs for a GRM item: SE(a), SE(d1), SE(d2), SE(d3)
cov_mix$se[["GRM1"]]
```

---

## `score_resp()` — Score Selected-Response Item Data

The functions above all start from item parameters that are either known
(`drm()`, `prm()`, `info()`, `traceline()`, `covirt()`) or estimated from an
already-scored response matrix. In practice, raw testing data usually looks
different: for each examinee, each item records the option that was actually
chosen (e.g., `2`, `"C"`), not yet a 0/1 correct/incorrect indicator.
`score_resp()` bridges that gap. Given the raw option choices and an answer
key, it returns a dichotomously scored (0/1) item-response matrix, of exactly
the kind consumed by `est_irt()`, `est_score()`, and the package's classical
test theory (CTT) functions introduced in the
*Classical Test Theory (CTT) Analysis* article.

`score_resp()` infers each item's option-coding scheme independently from its
own answer-key value, so a single test form may freely mix items that use
different coding conventions:

- a **numeric** key (e.g., `4`) is compared to responses as a number,
- a **Latin-letter** key (e.g., `"D"`) is compared case-insensitively as a
  letter (`"d"` matches `"D"`),
- any other non-blank key (e.g., a Korean syllable label, or a Roman numeral
  written in a non-Latin numeral script) is treated as a **general label** and
  matched with a case-insensitive exact string comparison.

An omitted response and a double-marked response (e.g., `"1,5"` for a
selected-response item where the examinee marked two options) are both scored
as incorrect (0), but their counts are tabulated separately in the returned
`resp_summary` so that omission and double-marking rates can be reported on
their own.

```{r score-resp-numeric}
# 5 examinees, 3 items, coded with option numbers; item 2 has one
# double-marked response ("2,4") and item 3 has one blank (NA) response
raw_num <- data.frame(
  V1 = c("1", "2", "3", "1", "2"),
  V2 = c("4", "4", "2,4", "4", "1"),
  V3 = c("2", "1", "3", "3", NA)
)
key_num <- c(1, 4, 3)   # correct option for V1, V2, V3
out_num <- score_resp(data = raw_num, key = key_num)

out_num$scored          # 0/1 scored matrix, same dimensions as raw_num
out_num$resp_summary    # one row per item: n_correct, n_blank, pct_double, ...
```

```{r score-resp-letters}
# Same idea, but items are coded with letter option labels instead of
# numbers; a lowercase response ("c") is still matched to key "C"
raw_let <- data.frame(
  V1 = c("A", "B", "C", "A", "B"),
  V2 = c("D", "D", "B,D", "D", "A"),
  V3 = c("B", "A", "C", "c", NA)
)
key_let <- c("A", "D", "C")
out_let <- score_resp(data = raw_let, key = key_let)

out_let$scored
identical(out_let$scored, out_num$scored)  # same scoring pattern as above
```

Because the coding scheme is detected per item, a numerically coded item and
a letter-coded item can appear side by side in the same call to
`score_resp()`, with each item's own key value deciding how that item's
column is interpreted:

```{r score-resp-mixed}
# Item 1 is numerically coded; items 2-3 are letter-coded
raw_mix <- data.frame(
  V1 = c("1", "2", "1", "1", "3"),
  V2 = c("D", "D", "D", "B,D", "D"),
  V3 = c("B", "A", "C", "C", NA)
)
key_mix <- c(1, "D", "C")
out_mix <- score_resp(data = raw_mix, key = key_mix)
out_mix$scored
```

Option labels written in scripts other than Latin letters (Korean syllables,
Roman numerals rendered with non-Latin numeral characters, and so on) are
supported as well, under the general-label rule described above; see
`?score_resp` for a worked example
using Korean syllable labels.

---

## Summary

| Function       | Input             | Key output                           | Used by |
|----------------|-------------------|--------------------------------------|---------|
| `simdat()`     | Item metadata or raw params + θ | Response matrix       | Testing, demos |
| `drm()`        | a, b, g, θ        | P(correct) matrix (θ × items)        | `simdat()`, `traceline()`, `info()`, `covirt()` |
| `prm()`        | a, d, θ, model    | Category prob. matrix (θ × cats)     | `simdat()`, `traceline()`, `info()`, `covirt()` |
| `info()`       | Item metadata + θ grid | IIF matrix, TIF vector; `plot()` | `est_score()` |
| `traceline()`  | Item metadata + θ grid | ICC matrix, TCC vector, category probs; `plot()` | Visualisation |
| `lwrc()`       | Item metadata + θ, or prob matrix | $\Pr(X = s \mid \theta)$ | `cac_lee()`, `sx2_fit()`, `est_score()` |
| `gen.weight()` | Distribution spec | Node–weight data frame               | `cac_lee()`, `cac_rud()`, `est_score()`, `covirt()` |
| `covirt()`     | Item metadata + sample size | Cov matrices, ASE vectors | SE approximation |
| `score_resp()` | Raw option choices + answer key | 0/1 scored response matrix | `est_irt()`, `est_score()`, `ctt()`, `ctt_distr()` |

---

## References
